---
title: "Available Curves in tcplfit2"
author: "Center for Computational Toxicology and Exposure"
output:
   prettydoc::html_pretty:
    theme: architect
    toc: yes
    toc_depth: 4
vignette: >
  %\VignetteIndexEntry{Available Curves in tcplfit2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(tcplfit2)
library(ggplot2)

data("mc3")
ex_conc <- seq(10^(min(mc3[,"logc"])),10^(max(mc3[,"logc"])),length.out = 100)
```

This vignette contains model details for all models available in `tcplfit2` package with parameter explanations and plots.  

# Continuous Response Model

## Exponential 2
Model: $f(x) = a*(e^{\frac{x}{b}}-1)$

Parameters: 

* `a` : The y-scale. If `a` increases, the curve is stretched vertically and if `a` decreases, the curve is shrunk vertically.
* `b` : The x-scale. If `b` increases, the curve is shrunk horizontally; if `b` decreases, the curve is stretched horizontally. 

```{r exp2, echo=FALSE, fig.width=5, fig.height=5}
fits_exp2 <- data.frame(
  # base line 
  y1 = exp2(ps = c(a = 0.45,b = 12,er = 0.1), x = ex_conc),
  
  # change a
  y2 = exp2(ps = c(a = 1.5,b = 12,er = 0.1), x = ex_conc),
  y3 = exp2(ps = c(a = 0.1,b = 12,er = 0.1), x = ex_conc),
  
  # change b
  y4 = exp2(ps = c(a = 0.45,b = 13.5,er = 0.1), x = ex_conc),
  y5 = exp2(ps = c(a = 0.45,b = 9,er = 0.1), x = ex_conc)
  
)

exp2_plot1 <- ggplot(fits_exp2, aes(ex_conc)) +  
    geom_line(aes(y = y1, color = "a=0.45")) +
     geom_line(aes(y = y2, color = "a=0.6")) +
   geom_line(aes(y = y3, color = "a=0.1")) +
  labs(x = "x", y = "y") +
  theme(legend.position = c(0.1,0.8)) +
  scale_color_manual(name='a values',
                     breaks=c('a=0.45', 'a=0.6', 'a=0.1'),
                     values=c('a=0.45'='black', 'a=0.6'='red', 'a=0.1'='green'))

exp2_plot2 <- ggplot(fits_exp2, aes(ex_conc)) +  
    geom_line(aes(y = y1, color = "b=12")) +
     geom_line(aes(y = y4, color = "b=13.5")) +
   geom_line(aes(y = y5, color = "b=9")) +
  labs(x = "x", y = "y") +
  theme(legend.position = c(0.1,0.8)) +
  scale_color_manual(name='b values',
                     breaks=c('b=12', 'b=13.5', 'b=9'),
                     values=c('b=12'='black', 'b=13.5'='red', 'b=9'='green'))


exp2_plot1
exp2_plot2

```

## Exponential 3
Model: $f(x) = a*(e^{(x/b)^p} - 1)$

The definitions of `a` and `b` are similar to those in Exponential 2, but it has one additional parameter `p`.

* `p` : Power. A measure of how steep the curve is. The larger `p` is the steeper the curve is.

```{r exp3, echo=FALSE, fig.width=5, fig.height=5}
fits_exp3 <- data.frame(
  # base line 
  y1 = exp3(ps = c(a = 1.67,b = 12.5,p = 0.87,er = 0.1), x = ex_conc),
  
  # change a
  y2 = exp3(ps = c(a = 0.8,b = 12.5,p = 0.87,er = 0.1), x = ex_conc),
  y3 = exp3(ps = c(a = 2.5,b = 12.5,p = 0.87,er = 0.1), x = ex_conc),
  
  # change b
  y4 = exp3(ps = c(a = 1.67,b = 11,p = 0.87,er = 0.1), x = ex_conc),
  y5 = exp3(ps = c(a = 1.67,b = 14,p = 0.87,er = 0.1), x = ex_conc),
  
  # change p
  y6 = exp3(ps = c(a = 1.67,b = 12.5,p = 0.9,er = 0.1), x = ex_conc),
  y7 = exp3(ps = c(a = 1.67,b = 12.5,p = 0.7,er = 0.1), x = ex_conc)
  
)

exp3_plot1 <- ggplot(fits_exp3, aes(ex_conc)) +  
    geom_line(aes(y = y1, color = "a=1.67")) +
     geom_line(aes(y = y2, color = "a=0.8")) +
   geom_line(aes(y = y3, color = "a=2.5")) +
  labs(x = "x", y = "y") +
  theme(legend.position = c(0.1,0.8)) +
  scale_color_manual(name='a values',
                     breaks=c('a=1.67', 'a=0.8', 'b=9'),
                     values=c('a=1.67'='black', 'a=0.8' = 'red', 'a=2.5'='green'))

exp3_plot2 <- ggplot(fits_exp3, aes(ex_conc)) +  
    geom_line(aes(y = y1, color = "b=12.5")) +
     geom_line(aes(y = y4, color = "b=11")) +
   geom_line(aes(y = y5, color = "b=14")) +
  labs(x = "x", y = "y") +
  theme(legend.position = c(0.1,0.8)) +
  scale_color_manual(name='b values',
                     breaks=c('b=12.5', 'b=11', 'b=14'),
                     values=c('b=12.5'='black', 'b=11'='red', 'b=14'='green'))

exp3_plot3 <- ggplot(fits_exp3, aes(ex_conc)) +  
    geom_line(aes(y = y1, color = "p=0.87")) +
     geom_line(aes(y = y6, color = "p=0.9")) +
   geom_line(aes(y = y7, color = "p=0.7")) +
  labs(x = "x", y = "y") +
  theme(legend.position = c(0.1,0.8)) +
  scale_color_manual(name='p values',
                     breaks=c('p=0.87', 'p=0.9', 'p=0.7'),
                     values=c('p=0.87'='black', 'p=0.9'='red', 'p=0.7'='green'))

#exp3_plot1
#exp3_plot2
exp3_plot3
```

## Exponential 4
Model: $f(x) = tp*(1-2^{(-\frac{x}{ga})})$

Parameters: 

* `tp` : Top. The horizontal asymptote the curve is approaching (also can be negative); it is the maximum or minimum of the predicted responses. 
* `ga` : AC50. Concentration at 50% of maximal activity, acts as slope for x and controls how fast the curve is approaching the top. If `ga` increases, the curve is shrunk horizontally; if `ga` decrease, the curve is stretched horizontally. 

```{r exp4, echo=FALSE, fig.width=5, fig.height=5}
fits_exp4 <- data.frame(
  # base line 
  y1 = exp4(ps = c(tp = 895,ga = 15,er = 0.1),x = ex_conc),
  
  # change tp
  y2 = exp4(ps = c(tp = 950,ga = 15,er = 0.1),x = ex_conc),
  y3 = exp4(ps = c(tp = 700,ga = 15,er = 0.1),x = ex_conc),
  
  # change ga
  y4 = exp4(ps = c(tp = 895,ga = 18,er = 0.1),x = ex_conc),
  y5 = exp4(ps = c(tp = 895,ga = 10,er = 0.1),x = ex_conc)
  
)

exp4_plot1 <- ggplot(fits_exp4, aes(ex_conc)) +  
    geom_line(aes(y = y1, color = "tp=895")) +
     geom_line(aes(y = y2, color = "tp=950")) +
   geom_line(aes(y = y3, color = "tp=700")) +
  labs(x = "x", y = "y") +
  theme(legend.position = c(0.1,0.8)) +
  scale_color_manual(name='tp values',
                     breaks=c('tp=895', 'tp=950', 'tp=700'),
                     values=c('tp=895'='black', 'tp=950'='red', 'tp=700'='green'))

exp4_plot2 <- ggplot(fits_exp4, aes(ex_conc)) +  
    geom_line(aes(y = y1, color = "ga=15")) +
     geom_line(aes(y = y4, color = "ga=18")) +
   geom_line(aes(y = y5, color = "ga=10")) +
  labs(x = "x", y = "y") +
  theme(legend.position = c(0.1,0.8)) +
  scale_color_manual(name='ga values',
                     breaks=c('ga=15', 'ga=18', 'ga=10'),
                     values=c('ga=15'='black', 'ga=18'='red', 'ga=10'='green'))


exp4_plot1
exp4_plot2
```


## Exponential 5
Model: $f(x) = tp*(1-2^{(-(x/ga)^p)})$

The definitions of `tp` and `ga` are similar to those in Exponential 4, and this has one additional parameter `p`. 

* `p` : Power. A measure of how steep the curve is. The larger `p` is the steeper the curve is.

```{r exp5, echo=FALSE, fig.width=5, fig.height=5}
fits_exp5 <- data.frame(
  # base line 
  y1 = exp5(ps = c(tp = 793,ga = 6.25,p = 1.25,er = 0.1), x = ex_conc),
  
  # change tp
  y2 = exp5(ps = c(tp = 900,ga = 6.25,p = 1.25,er = 0.1), x = ex_conc),
  y3 = exp5(ps = c(tp = 500,ga = 6.25,p = 1.25,er = 0.1), x = ex_conc),
  
  # change ga
  y4 = exp5(ps = c(tp = 793,ga = 8.2,p = 1.25,er = 0.1), x = ex_conc),
  y5 = exp5(ps = c(tp = 793,ga = 6,p = 1.25,er = 0.1), x = ex_conc),
  
  # change p
  y6 = exp5(ps = c(tp = 793,ga = 6.25,p = 0.97,er = 0.1), x = ex_conc),
  y7 = exp5(ps = c(tp = 793,ga = 6.25,p = 1.5,er = 0.1), x = ex_conc)
  
)

exp5_plot1 <- ggplot(fits_exp5, aes(ex_conc)) +  
    geom_line(aes(y = y1, color = "tp=793")) +
     geom_line(aes(y = y2, color = "tp=900")) +
   geom_line(aes(y = y3, color = "tp=500")) +
  labs(x = "x", y = "y") +
  theme(legend.position = c(0.1,0.8)) +
  scale_color_manual(name='tp values',
                     breaks=c('tp=793', 'tp=900', 'tp=500'),
                     values=c('tp=793'='black', 'tp=900' = 'red', 'tp=500'='green'))

exp5_plot2 <- ggplot(fits_exp5, aes(ex_conc)) +  
    geom_line(aes(y = y1, color = "ga=6.25")) +
     geom_line(aes(y = y4, color = "ga=8.2")) +
   geom_line(aes(y = y5, color = "ga=6")) +
  labs(x = "x", y = "y") +
  theme(legend.position = c(0.1,0.8)) +
  scale_color_manual(name='ga values',
                     breaks=c('ga=6.25', 'ga=8.2', 'ga=6'),
                     values=c('ga=6.25'='black', 'ga=8.2'='red', 'ga=6'='green'))

exp5_plot3 <- ggplot(fits_exp5, aes(ex_conc)) +  
    geom_line(aes(y = y1, color = "p=1.25")) +
     geom_line(aes(y = y6, color = "p=0.99")) +
   geom_line(aes(y = y7, color = "p=1.5")) +
  labs(x = "x", y = "y") +
  theme(legend.position = c(0.1,0.8)) +
  scale_color_manual(name='p values',
                     breaks=c('p=1.25', 'p=0.97', 'p=1.5'),
                     values=c('p=1.25'='black', 'p=0.97'='red', 'p=1.5'='green'))

#exp5_plot1
#exp5_plot2
exp5_plot3
```


## Polynomial 1 & 2

Polynomial 1 is linear line $y = ax$ where `a` is the slope of the line. 

Polynomial 2 is a parabola and this package has slightly different parameterzation from the normal quadratic polynomial equation. 

The usual quadratic polynomial is $f(x) = ax^2+bx+c$. In this package is $f(x) = a(\frac{x}{b} + \frac{x^2}{b^2})$. 

Parameters: 

* `a` : The y-scale. The sign of this parameter determines if the vertex will land below or below x-axis. If `a` increases, the curve is stretched vertically. If `a` decreases, the curve is shrunk vertically.

* `b` : The x-scale. If `b` increase, the curve is shrunk horizontally, and if `b` decrease, the curve is stretched horizontally. `b` is restricted to be non-negative, so the sign of `a` determines the directionality of the curve. If it is a decreasing curve, `a` will be negative. 

```{r poly, echo=FALSE, fig.width=5, fig.height=5}
fits_poly <- data.frame(
  # base line 
  y1 = poly1(ps = c(a = 3.5,er = 0.1),x = ex_conc),
  y1.2 = poly2(ps = c(a = 0.13,b = 2,er = 0.1),x = ex_conc),
  
  # change a for poly1
  y2 = poly1(ps = c(a = 5,er = 0.1),x = ex_conc),
  y3 = poly1(ps = c(a = 2.6,er = 0.1),x = ex_conc),

  # change a for poly2
  y4 = poly2(ps = c(a = 1, b = 2,er = 0.1),x = ex_conc),
  y5 = poly2(ps = c(a = 0.1, b = 2,er = 0.1),x = ex_conc),
  
  # change b for poly2
  y6 = poly2(ps = c(a = 0.13,b = 4,er = 0.1),x = ex_conc),
  y7 = poly2(ps = c(a = 0.13,b = 0.5,er = 0.1),x = ex_conc)
  
)

poly1_plot <- ggplot(fits_poly, aes(ex_conc)) +  
     geom_line(aes(y = y1, color = "a=3.5")) +
     geom_line(aes(y = y2, color = "a=5")) +
     geom_line(aes(y = y3, color = "a=2.6")) +
  labs(x = "x", y = "y") +
  theme(legend.position = c(0.1,0.8)) +
  scale_color_manual(name='a values',
                     breaks=c('a=3.5', 'a=5', 'a=2.6'),
                     values=c('a=3.5'='black', 'a=5' = 'red', 'a=2.6'='green'))

poly2_plot1 <- ggplot(fits_poly, aes(ex_conc)) +  
    geom_line(aes(y = y1.2, color = "a=0.13")) +
     geom_line(aes(y = y4, color = "a=1")) +
   geom_line(aes(y = y5, color = "a=0.1")) +
  labs(x = "x", y = "y") +
  theme(legend.position = c(0.1,0.8)) +
  scale_color_manual(name='a values',
                     breaks=c('a=0.13', 'a=1', 'a=0.1'),
                     values=c('a=0.13'='black', 'a=1'='red', 'a=0.1'='green'))

poly2_plot2 <- ggplot(fits_poly, aes(ex_conc)) +  
    geom_line(aes(y = y1.2, color = "b=2")) +
     geom_line(aes(y = y6, color = "b=4")) +
   geom_line(aes(y = y7, color = "b=0.5")) +
  labs(x = "x", y = "y") +
  theme(legend.position = c(0.1,0.8)) +
  scale_color_manual(name='b values',
                     breaks=c('b=2', 'b=4', 'b=0.5'),
                     values=c('b=2'='black', 'b=4'='red', 'b=0.5'='green'))

#poly1_plot
poly2_plot1
poly2_plot2
```

## Power 
Model: $f(x) = a*x^b$

Parameters: 

* `a` : The scaling factor. If `a` increases, the curve is stretched vertically. If `a` decreases, the curve is shrunk vertically.
* `p` : Power, or the rate of growth. A measure of how steep the curve is. The larger `p` is the steeper the curve is.

```{r pow, echo=FALSE, fig.width=5, fig.height=5}
fits_pow <- data.frame(
  # base line 
  y1 = pow(ps = c(a = 1.23,p = 1.45,er = 0.1),x = ex_conc),
  
  # change a
  y2 = pow(ps = c(a = 3.2,p = 1.45,er = 0.1),x = ex_conc),
  y3 = pow(ps = c(a = 0.8,p = 1.45,er = 0.1),x = ex_conc),
  
  # change p
  y4 = pow(ps = c(a = 1.23,p = 1.5,er = 0.1),x = ex_conc),
  y5 = pow(ps = c(a = 1.23,p = 1.2,er = 0.1),x = ex_conc)
  
)

pow_plot1 <- ggplot(fits_pow, aes(ex_conc)) +  
     geom_line(aes(y = y1, color = "a=1.23")) +
     geom_line(aes(y = y2, color = "a=3.2")) +
     geom_line(aes(y = y3, color = "a=0.8")) +
  labs(x = "x", y = "y") +
  theme(legend.position = c(0.1,0.8)) +
  scale_color_manual(name='a values',
                     breaks=c('a=1.23', 'a=3.2', 'a=0.8'),
                     values=c('a=1.23'='black', 'a=3.2'='red', 'a=0.8'='green'))

pow_plot2 <- ggplot(fits_pow, aes(ex_conc)) +  
    geom_line(aes(y = y1, color = "p=1.45")) +
     geom_line(aes(y = y4, color = "p=1.5")) +
   geom_line(aes(y = y5, color = "p=1.2")) +
  labs(x = "x", y = "y") +
  theme(legend.position = c(0.1,0.8)) +
  scale_color_manual(name='p values',
                     breaks=c('p=1.45', 'p=1.5', 'p=1.2'),
                     values=c('p=1.45'='black', 'p=1.5'='red', 'p=1.2'='green'))


pow_plot1
pow_plot2
```


## Hill
Model: $\frac{tp}{(1 + (ga/x)^p )}$

Parameters: 

* `tp` : Top, the highest or lowest response achieved at saturation. It is the horizontal asymptote the curve is approaching. 
* `ga` : AC50. Concentration at 50% of maximal activity, it provides useful information about the "apparent affinity" of the protein under study (enzyme, transporter, etc.) for the substrate.
* `p` : Hill coefficient, acts as power, a measure of how steep the response curve is. It's a measure of the cooperativity of substrate binding to the enzyme, transporter, etc.

```{r hill, echo=FALSE, fig.height=5, fig.width=5}
fits_hill <- data.frame(
  # base line 
  y1 = hillfn(ps = c(tp = 750,ga = 5,p = 1.76,er = 0.1), x = ex_conc),
  
  # change tp
  y2 = hillfn(ps = c(tp = 850,ga = 5,p = 1.76,er = 0.1), x = ex_conc),

  # change ga
  y3 = hillfn(ps = c(tp = 750,ga = 7,p = 1.76,er = 0.1), x = ex_conc),
  y4 = hillfn(ps = c(tp = 750,ga = 4,p = 1.76,er = 0.1), x = ex_conc),
  
  # change p
  y5 = hillfn(ps = c(tp = 750,ga = 5,p = 2,er = 0.1), x = ex_conc),
  y6 = hillfn(ps = c(tp = 750,ga = 5,p = 1.5,er = 0.1), x = ex_conc)
  
)

hill_plot1 <- ggplot(fits_hill, aes(ex_conc)) +  
    geom_line(aes(y = y1, color = "ga=5")) +
     geom_line(aes(y = y3, color = "ga=7")) +
   geom_line(aes(y = y4, color = "ga=4")) +
  labs(x = "x", y = "y") +
  theme(legend.position = c(0.8,0.1)) +
  scale_color_manual(name='ga values',
                     breaks=c('ga=5', 'ga=7', 'ga=4'),
                     values=c('ga=5'='black', 'ga=7'='red', 'ga=4'='green'))


hill_plot2 <- ggplot(fits_hill, aes(ex_conc)) +  
    geom_line(aes(y = y1, color = "p=1.76")) +
     geom_line(aes(y = y4, color = "p=2")) +
   geom_line(aes(y = y5, color = "p=1.5")) +
  labs(x = "x", y = "y") +
  theme(legend.position = c(0.8,0.1)) +
  scale_color_manual(name='p values',
                     breaks=c('p=1.76', 'p=2', 'p=1.5'),
                     values=c('p=1.76'='black', 'p=2'='red', 'p=1.5'='green'))

hill_plot1
hill_plot2
```


## Gain-Loss
Model: $\frac{tp}{[(1 + (ga/x)^p )(1 + (x/la)^q )]}$

Gain-Loss is the product of two hill functions, one simulates the response going up, one simulates the responses going down. A curve can gain first and then loss or the opposite. 

The definitions of `tp`, `ga`, and `p` are the same as in hill model, and `la` is the counterpart of `ga`, `q` is the counterpart of `p` in the decreasing half of the curve. 

* `la` : Loss AC50. The larger it gets the slower the curve starts to decrease (if it increase first). 
* `q` : Loss power or the rate of loss. The larger it is the faster the curve decreases 

```{r gnls, echo=FALSE, fig.width=5, fig.height=5}
fits_gnls <- data.frame(
  # base line 
  y1 = gnls(ps = c(tp = 750,ga = 15,p = 1.45,la = 50,q = 1.34, er = 0.1), x = ex_conc),
  
  # change la
  y2 = gnls(ps = c(tp = 750,ga = 15,p = 1.45,la = 70,q = 1.34, er = 0.1), x = ex_conc),
  y3 = gnls(ps = c(tp = 750,ga = 15,p = 1.45,la = 20,q = 1.34, er = 0.1), x = ex_conc),
  
  # change q
  y4 = gnls(ps = c(tp = 750,ga = 15,p = 1.45,la = 50,q = 1, er = 0.1), x = ex_conc),
  y5 = gnls(ps = c(tp = 750,ga = 15,p = 1.45,la = 50,q = 1.5, er = 0.1), x = ex_conc)
  
)

gnls_plot1 <- ggplot(fits_gnls, aes(log10(ex_conc))) +  
  geom_line(aes(y = y1, color = "la=50")) +
     geom_line(aes(y = y2, color = "la=70")) +
   geom_line(aes(y = y3, color = "la=20")) +
  labs(x = "x", y = "y") +
  theme(legend.position = c(0.1,0.8)) +
  scale_color_manual(name='la values',
                     breaks=c('la=50', 'la=70', 'la=20'),
                     values=c('la=50'='black', 'la=70'='red', 'la=20'='green'))

gnls_plot2 <- ggplot(fits_gnls, aes(log10(ex_conc))) +  
  geom_line(aes(y = y1, color = "q=1.34")) +
     geom_line(aes(y = y4, color = "q=1")) +
   geom_line(aes(y = y5, color = "q=1.5")) +
  labs(x = "x", y = "y") +
  theme(legend.position = c(0.1,0.8)) +
  scale_color_manual(name='q values',
                     breaks=c('q=1.34', 'q=1', 'q=1.5'),
                     values=c('q=1.34'='black', 'q=1'='red', 'q=1.5'='green'))
  
  
gnls_plot1
gnls_plot2
```



