---
title: "Details for Available Models in tcplfit2"
author: "Center for Computational Toxicology and Exposure"
output:
   prettydoc::html_pretty:
    theme: architect
    toc: yes
    toc_depth: 4
vignette: >
  %\VignetteIndexEntry{Details_for_Available_Models_in_tcplfit2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{css, echo=FALSE}
.scroll-100 {
  max-height: 100px;
  overflow-y: auto;
}
```

```{r setup}
library(tcplfit2)
library(ggplot2)
library(gridExtra)

# prepare data for demonstration
data("mc3")
ex_conc <- seq(10^(min(mc3[,"logc"])),10^(max(mc3[,"logc"])),length.out = 100)
```

This vignette contains model details for all models available in `tcplfit2` package with parameter explanations and plots.  

# Continuous Response Model

## Polynomial 1

Polynomial 1 is a simple linear model $y = ax$ where `a` is the slope of the line (i.e. rate of change for the response across the dose range). If bi-directional fitting is allowed, then $-\infty < a <\infty$. Otherwise, $a \ge 0$ (i.e. non-negative).

```{r poly 1, class.source="scroll-100", fig.width=8, fig.height=5}
poly1_plot <- ggplot(mapping=aes(ex_conc)) +  
  geom_line(aes(y = 55*ex_conc, color = "a=55")) +
  geom_line(aes(y = 10*ex_conc, color = "a=10")) +
  geom_line(aes(y = 0.05*ex_conc, color = "a=0.05")) +
  geom_line(aes(y = -5*ex_conc, color = "a=-5")) +
  labs(x = "x", y = "y") +
  theme(legend.position = c(0.1,0.8)) +
  scale_color_manual(name='a values',
                     breaks=c('a=-5', 'a=0.05', 'a=10', 'a=55'),
                     values=c('a=-5'='black', 'a=0.05' = 'red', 'a=10'='blue', 'a=55'='yellow'))

poly1_plot
```

**Figure 1:** This plot hows how changes in slope `a` affect the lines.


## Polynomial 2

Polynomial 2 is a quadratic model fit. Note, the version implemented in `tcplfit2` is reparametrized to assume the baseline response is always zero, and the `a` and `b` parameters are interpreted in terms of their impact on the the x- and y-scale, respectively.

* `tcplfit2`: $f(x) = a(\frac{x}{b} + \frac{x^2}{b^2})$. 
* typical polynomial 2: $f(x) = ax^2+bx+c$. 

Parameters: 

* `a` : The y-scale. If `a` increases, the curve is stretched vertically. If bi-directional fitting is allowed, then $-\infty < a <\infty $. Otherwise, $a \ge 0$ (i.e. non-negative).

* `b` : The x-scale. If `b` increase, the curve is shrunk horizontally. Polynomial 2 model is optimized such that $b > 0$. 

```{r poly 2, class.source="scroll-100", fig.width=8, fig.height=5}
fits_poly <- data.frame(
  # change a 
  y1 = poly2(ps = c(a = 40, b = 2),x = ex_conc),
  y2 = poly2(ps = c(a = 6, b = 2),x = ex_conc),
  y3 = poly2(ps = c(a = 0.1, b = 2),x = ex_conc),
  y4 = poly2(ps = c(a = -2, b = 2),x = ex_conc),
  y5 = poly2(ps = c(a = -20, b = 2),x = ex_conc),
  
  # change b 
  y6 = poly2(ps = c(a = 4,b = 1.8),x = ex_conc),
  y7 = poly2(ps = c(a = 4,b = 7),x = ex_conc),
  y8 = poly2(ps = c(a = 4,b = 16),x = ex_conc)
)

# shows how changes in parameter 'a' affect the shape of the curve 
poly2_plot1 <- ggplot(fits_poly, aes(ex_conc)) +
  geom_line(aes(y = y1, color = "a=40")) +
  geom_line(aes(y = y2, color = "a=6")) +
  geom_line(aes(y = y3, color = "a=0.1")) +
  geom_line(aes(y = y4, color = "a=-2")) +
  geom_line(aes(y = y5, color = "a=-20")) +
  labs(x = "x", y = "y") +
  theme(legend.position = c(0.15,0.8)) +
  scale_color_manual(name='a values',
                     breaks=c('a=-20', 'a=-2', 'a=0.1', 'a=6', 'a=40'),
                     values=c('a=-20'='black', 'a=-2'='red', 'a=0.1'='blue', 'a=6'='yellow', 'a=40'='grey'))

# shows how changes in parameter 'b' affect the shape of the curve 
poly2_plot2 <- ggplot(fits_poly, aes(ex_conc)) +  
  geom_line(aes(y = y6, color = "b=1.8")) +
  geom_line(aes(y = y7, color = "b=7")) +
  geom_line(aes(y = y8, color = "b=16")) +
  labs(x = "x", y = "y") +
  theme(legend.position = c(0.15,0.8)) +
  scale_color_manual(name='b values',
                     breaks=c('b=1.8', 'b=7', 'b=16'),
                     values=c('b=1.8'='black', 'b=7'='red', 'b=16'='blue'))

grid.arrange(poly2_plot1, poly2_plot2, ncol = 2)
```

**Figure 2:** Plot on the left shows how changes in parameter `a` affect the shape of polynomial 2 curves, while holding `b` constant. Plot on the right shows how changes in parameter `b` affect the shape of polynomial 2 curves, while holding `a` constant.

## Power 
Model: $f(x) = a*x^b$

Parameters: 

* `a` : The scaling factor. If `a` increases, the curve is stretched vertically. If bi-directional fitting is allowed, then $-\infty < a <\infty $. Otherwise, $a \gt 0$.
* `p` : Power, or the rate of growth. A measure of how steep the curve is. The larger `p` is the steeper the curve is. `p` is restricted such that $0.3 \le p \le 20$. 

```{r pow, class.source="scroll-100", fig.width=8, fig.height=5}
fits_pow <- data.frame(
  # change a
  y1 = pow(ps = c(a = 0.48,p = 1.45),x = ex_conc),
  y2 = pow(ps = c(a = 7.2,p = 1.45),x = ex_conc),
  y3 = pow(ps = c(a = -3.2,p = 1.45),x = ex_conc),
  
  # change p
  y4 = pow(ps = c(a = 1.2,p = 0.3),x = ex_conc),
  y5 = pow(ps = c(a = 1.2,p = 1.6),x = ex_conc),
  y6 = pow(ps = c(a = 1.2,p = 2),x = ex_conc)
)

# shows how changes in parameter 'a' affect the shape of the curve
pow_plot1 <- ggplot(fits_pow, aes(ex_conc)) +  
  geom_line(aes(y = y1, color = "a=0.48")) +
  geom_line(aes(y = y2, color = "a=7.2")) +
  geom_line(aes(y = y3, color = "a=-3.2")) +
  labs(x = "x", y = "y") +
  theme(legend.position = c(0.15,0.8)) +
  scale_color_manual(name='a values',
                     breaks=c('a=-3.2', 'a=0.48', 'a=7.2'),
                     values=c('a=-3.2'='black', 'a=0.48'='red', 'a=7.2'='blue'))

# shows how changes in parameter 'p' affect the shape of the curve
pow_plot2 <- ggplot(fits_pow, aes(ex_conc)) +  
  geom_line(aes(y = y4, color = "p=0.3")) +
  geom_line(aes(y = y5, color = "p=1.6")) +
  geom_line(aes(y = y6, color = "p=2")) +
  labs(x = "x", y = "y") +
  theme(legend.position = c(0.15,0.8)) +
  scale_color_manual(name='p values',
                     breaks=c('p=0.3', 'p=1.6', 'p=2'),
                     values=c('p=0.3'='black', 'p=1.6'='red', 'p=2'='blue'))



grid.arrange(pow_plot1, pow_plot2, ncol = 2)
```

**Figure 3:** Plot on the left shows how changes in parameter `a` affect the shape of power curves, while holding `p` constant. Plot on the right shows how changes in parameter `p` affect the shape of power curves, while holding `a` constant.


## Hill {#hill}
Model: $f(x) = \frac{tp}{(1 + (ga/x)^p )}$

Parameters: 

* `tp` : Top, the highest or lowest response achieved at saturation. It is the horizontal asymptote the curve is approaching. If bi-directional fitting is allowed, then $-\infty < tp <\infty$. Otherwise $0 \le tp < \infty$.
* `ga` : AC50. Concentration at 50% of maximal activity, it provides useful information about the "apparent affinity" of the protein under study (enzyme, transporter, etc.) for the substrate. The parameter constrain on `ga` is $-\infty <ga < \infty$ 
* `p` : Hill coefficient, acts as power, a measure of how steep the response curve is. It's a measure of the cooperativity of substrate binding to the enzyme, transporter, etc. `p` is restricted such that $0.3 \le p \le 8$. 

```{r hill, class.source="scroll-100", fig.height=5, fig.width=8}
fits_hill <- data.frame(
  # change tp
  y1 = hillfn(ps = c(tp = 750,ga = 5,p = 1.76), x = ex_conc),
  y2 = hillfn(ps = c(tp = 850,ga = 5,p = 1.76), x = ex_conc),

  # change ga
  y3 = hillfn(ps = c(tp = 750,ga = 7,p = 1.76), x = ex_conc),
  y4 = hillfn(ps = c(tp = 750,ga = 4,p = 1.76), x = ex_conc),
  
  # change p
  y5 = hillfn(ps = c(tp = 750,ga = 5,p = 2), x = ex_conc),
  y6 = hillfn(ps = c(tp = 750,ga = 5,p = 1.5), x = ex_conc)
  
)

# shows how changes in parameter 'ga' affect the shape of the curve
hill_plot1 <- ggplot(fits_hill, aes(ex_conc)) +  
    geom_line(aes(y = y1, color = "ga=5")) +
     geom_line(aes(y = y3, color = "ga=7")) +
   geom_line(aes(y = y4, color = "ga=4")) +
  labs(x = "x", y = "y") +
  theme(legend.position = c(0.8,0.1)) +
  scale_color_manual(name='ga values',
                     breaks=c('ga=5', 'ga=7', 'ga=4'),
                     values=c('ga=5'='black', 'ga=7'='red', 'ga=4'='blue'))

# shows how changes in parameter 'p' affect the shape of the curve
hill_plot2 <- ggplot(fits_hill, aes(ex_conc)) +  
    geom_line(aes(y = y1, color = "p=1.76")) +
     geom_line(aes(y = y5, color = "p=2")) +
   geom_line(aes(y = y6, color = "p=1.5")) +
  labs(x = "x", y = "y") +
  theme(legend.position = c(0.8,0.1)) +
  scale_color_manual(name='p values',
                     breaks=c('p=1.76', 'p=2', 'p=1.5'),
                     values=c('p=1.76'='black', 'p=2'='red', 'p=1.5'='blue'))


grid.arrange(hill_plot1, hill_plot1, ncol = 2)
```

**Figure 4:** Plot on the left shows how changes in parameter `ga` affect the shape of hill curves, while holding other parameters constant. Plot on the right shows how changes in parameter `p` affect the shape of power curves, while holding other parameters constant.


## Gain-Loss
Model: $f(x) = \frac{tp}{[(1 + (ga/x)^p )(1 + (x/la)^q )]}$

Gain-Loss is the product of two hill functions, one simulates the response going up, one simulates the responses going down. A curve can gain first and then loss or the opposite. 

The definitions of `tp`, `ga`, and `p` are the same as in [hill model](#hill), and `la` is the counterpart of `ga`, `q` is the counterpart of `p` in the decreasing half of the curve. 

* `la` : Loss AC50. The larger it gets the slower the curve starts to decrease (if it increase first). The parameter constrains on `la` are $-\infty <la < \infty$ and $la-ga\ge 1.5$.
* `q` : Loss power or the rate of loss. The larger it is the faster the curve decreases. `q` is restricted such that $0.3 \le p \le 8$. 

```{r gnls, class.source="scroll-100", fig.width=8, fig.height=5}
fits_gnls <- data.frame(
  # change la
  y1 = gnls(ps = c(tp = 750,ga = 15,p = 1.45,la = 17,q = 1.34), x = ex_conc),
  y2 = gnls(ps = c(tp = 750,ga = 15,p = 1.45,la = 50,q = 1.34), x = ex_conc),
  y3 = gnls(ps = c(tp = 750,ga = 15,p = 1.45,la = 100,q = 1.34), x = ex_conc),

  # change q
  y4 = gnls(ps = c(tp = 750,ga = 15,p = 1.45,la = 20,q = 0.3), x = ex_conc),
  y5 = gnls(ps = c(tp = 750,ga = 15,p = 1.45,la = 20,q = 1.2), x = ex_conc),
  y6 = gnls(ps = c(tp = 750,ga = 15,p = 1.45,la = 20,q = 8), x = ex_conc)
  
)

# shows how changes in parameter 'la' affect the shape of the curve
gnls_plot1 <- ggplot(fits_gnls, aes(log10(ex_conc))) +  
  geom_line(aes(y = y1, color = "la=17")) +
  geom_line(aes(y = y2, color = "la=50")) +
  geom_line(aes(y = y3, color = "la=100")) +
  labs(x = "x", y = "y") +
  theme(legend.position = c(0.1,0.8)) +
  scale_color_manual(name='la values',
                     breaks=c('la=17', 'la=50', 'la=100'),
                     values=c('la=17'='black', 'la=50'='red', 'la=100'='blue'))

# shows how changes in parameter 'q' affect the shape of the curve
gnls_plot2 <- ggplot(fits_gnls, aes(log10(ex_conc))) +  
  geom_line(aes(y = y4, color = "q=0.3")) +
  geom_line(aes(y = y5, color = "q=1.2")) +
  geom_line(aes(y = y6, color = "q=8")) +
  labs(x = "x", y = "y") +
  theme(legend.position = c(0.1,0.8)) +
  scale_color_manual(name='q values',
                     breaks=c('q=0.3', 'q=1.2', 'q=8'),
                     values=c('q=0.3'='black', 'q=1.2'='red', 'q=8'='blue'))
  
  
grid.arrange(gnls_plot1, gnls_plot2, ncol = 2)
```

**Figure 5:** Plot on the left shows how changes in parameter `la` affect the shape of Gain-Loss curves, while holding other parameters constant. Plot on the right shows how changes in parameter `q` affect the shape of power curves, while holding other parameters constant.

## Exponential 2 {#expoential2}
Model: $f(x) = a*(e^{\frac{x}{b}}-1)$

Parameters: 

* `a` : The y-scale. If `a` increases, the curve is stretched vertically. If bi-directional fitting is allowed, then $-\infty < a < \infty$. Otherwise, $0 < a <\infty$.
* `b` : The x-scale. If `b` increases, the curve is shrunk horizontally. `b` is restricted to $b > 0$ (i.e. positive). 

```{r exp2, class.source="scroll-100", fig.width=8, fig.height=5}
fits_exp2 <- data.frame(
  # change a
  y1 = exp2(ps = c(a = 20,b = 12), x = ex_conc),
  y2 = exp2(ps = c(a = 9,b = 12), x = ex_conc),
  y3 = exp2(ps = c(a = 0.1,b = 12), x = ex_conc),
  y4 = exp2(ps = c(a = -3,b = 12), x = ex_conc),
  
  # change b
  y5 = exp2(ps = c(a = 0.45,b = 4), x = ex_conc),
  y6 = exp2(ps = c(a = 0.45,b = 4.5), x = ex_conc),
  y7 = exp2(ps = c(a = 0.45,b = 20), x = ex_conc)
  
)

# shows how changes in parameter 'a' affect the shape of the curve 
exp2_plot1 <- ggplot(fits_exp2, aes(ex_conc)) +  
  geom_line(aes(y = y1, color = "a=20")) +
  geom_line(aes(y = y2, color = "a=9")) +
  geom_line(aes(y = y3, color = "a=0.1")) +
  geom_line(aes(y = y4, color = "a=-3")) +
  labs(x = "x", y = "y") +
  theme(legend.position = c(0.15,0.8)) +
  scale_color_manual(name='a values',
                     breaks=c('a=-3', 'a=0.1', 'a=9', 'a=20'),
                     values=c('a=-3'='black', 'a=0.1'='red', 'a=9'='blue', 'a=20'='yellow'))

# shows how changes in parameter 'b' affect the shape of the curve 
exp2_plot2 <- ggplot(fits_exp2, aes(ex_conc)) +  
  geom_line(aes(y = y5, color = "b=4")) +
  geom_line(aes(y = y6, color = "b=4.5")) +
  geom_line(aes(y = y7, color = "b=20")) +
  labs(x = "x", y = "y") +
  theme(legend.position = c(0.15,0.8)) +
  scale_color_manual(name='b values',
                     breaks=c('b=4', 'b=4.5', 'b=20'),
                     values=c('b=4'='black', 'b=4.5'='red', 'b=20'='blue'))

grid.arrange(exp2_plot1, exp2_plot2, ncol = 2)
```

**Figure 6:** Plot on the left shows how changes in parameter `a` affect the shape of exponential 2 curves, while holding `b` constant. Plot on the right shows how changes in parameter `b` affect the shape of exponential 2 curves, while holding `a` constant.

## Exponential 3
Model: $f(x) = a*(e^{(x/b)^p} - 1)$

The definitions of `a` and `b` are similar to those in Exponential 2, for details and plots please refer back to [Exponential 2](#expoential2). 

Exponential 3 has one additional parameter `p`.

* `p` : Power. A measure of how steep the curve is. The larger `p` is the steeper the curve is. `p` is restricted to be $0.3 \le p \le 8$. 

```{r exp3, class.source="scroll-100", fig.width=5, fig.height=5}
fits_exp3 <- data.frame(
  # change p
  y1 = exp3(ps = c(a = 1.67,b = 12.5,p = 0.3), x = ex_conc),
  y2 = exp3(ps = c(a = 1.67,b = 12.5,p = 0.9), x = ex_conc),
  y3 = exp3(ps = c(a = 1.67,b = 12.5,p = 1.2), x = ex_conc)
  
)

# shows how changes in parameter 'p' affect the shape of the curve 
exp3_plot <- ggplot(fits_exp3, aes(ex_conc)) +  
  geom_line(aes(y = y1, color = "p=0.87")) +
  geom_line(aes(y = y2, color = "p=0.9")) +
  geom_line(aes(y = y3, color = "p=0.7")) +
  labs(x = "x", y = "y") +
  theme(legend.position = c(0.15,0.8)) +
  scale_color_manual(name='p values',
                     breaks=c('p=0.87', 'p=0.9', 'p=0.7'),
                     values=c('p=0.87'='black', 'p=0.9'='red', 'p=0.7'='blue'))


exp3_plot
```

**Figure 7:** This plot shows how changes in parameter `p` affect the shape of exponential 3 curves, while holding other parameters constant.

## Exponential 4 {#expoential4}
Model: $f(x) = tp*(1-2^{(-\frac{x}{ga})})$

Parameters: 

* `tp` : Top. The horizontal asymptote the curve is approaching (also can be negative); it is the maximum or minimum of the predicted responses. If bi-directional fitting is allowed, then $-\infty <tp < \infty$. Otherwise, $0 \le tp < \infty$.
* `ga` : AC50. Concentration at 50% of maximal activity, acts as slope for x and controls how fast the curve is approaching the top. If `ga` increases, the curve is shrunk horizontally. `ga` is restricted to be $0 \le ga < \infty$ (e.g. non-negative). 

```{r exp4, class.source="scroll-100", fig.width=8, fig.height=5}
fits_exp4 <- data.frame(
  # change tp  
  y1 = exp4(ps = c(tp = 895,ga = 15),x = ex_conc),
  y2 = exp4(ps = c(tp = 200,ga = 15),x = ex_conc),
  y3 = exp4(ps = c(tp = -500,ga = 15),x = ex_conc),
  
  # change ga
  y4 = exp4(ps = c(tp = 500,ga = 0.4),x = ex_conc),
  y5 = exp4(ps = c(tp = 500,ga = 10),x = ex_conc),
  y6 = exp4(ps = c(tp = 500,ga = 20),x = ex_conc)
  
)

# shows how changes in parameter 'tp' affect the shape of the curve 
exp4_plot1 <- ggplot(fits_exp4, aes(ex_conc)) +  
  geom_line(aes(y = y1, color = "tp=895")) +
  geom_line(aes(y = y2, color = "tp=200")) +
  geom_line(aes(y = y3, color = "tp=-500")) +
  labs(x = "x", y = "y") +
  theme(legend.position = c(0.1,0.8)) +
  scale_color_manual(name='tp values',
                     breaks=c('tp=-500', 'tp=200', 'tp=895'),
                     values=c('tp=-500'='black', 'tp=200'='red', 'tp=895'='blue'))


# shows how changes in parameter 'ga' affect the shape of the curve 
exp4_plot2 <- ggplot(fits_exp4, aes(ex_conc)) +  
  geom_line(aes(y = y4, color = "ga=0.4")) +
  geom_line(aes(y = y5, color = "ga=10")) +
  geom_line(aes(y = y6, color = "ga=20")) +
  labs(x = "x", y = "y") +
  theme(legend.position = c(0.8,0.2)) +
  scale_color_manual(name='ga values',
                     breaks=c('ga=0.4', 'ga=10', 'ga=20'),
                     values=c('ga=0.4'='black', 'ga=10'='red', 'ga=20'='blue'))


grid.arrange(exp4_plot1, exp4_plot2, ncol = 2)
```

**Figure 8:** Plot on the left shows how changes in parameter `tp` affect the shape of exponential 4 curves, while holding `ga` constant. Plot on the right shows how changes in parameter `ga` affect the shape of exponential 2 curves, while holding `tp` constant.

## Exponential 5
Model: $f(x) = tp*(1-2^{(-(x/ga)^p)})$

The definitions of `tp` and `ga` are similar to those in Exponential 4, for details and plots please refer back to [Exponential 4](#expoential4). 

Exponential 5 has one additional parameter `p`. 

* `p` : Power. A measure of how steep the curve is. The larger `p` is the steeper the curve is. `p` is restricted to be $0.3 \le p \le 8$.

```{r exp5, class.source="scroll-100", fig.width=5, fig.height=5}
fits_exp5 <- data.frame(
  # change p
  y1 = exp5(ps = c(tp = 793,ga = 6.25,p = 0.3), x = ex_conc),
  y2 = exp5(ps = c(tp = 793,ga = 6.25,p = 3.4), x = ex_conc),
  y3 = exp5(ps = c(tp = 793,ga = 6.25,p = 8), x = ex_conc)
  
)

# shows how changes in parameter 'p' affect the shape of the curve 
exp5_plot <- ggplot(fits_exp5, aes(ex_conc)) +  
  geom_line(aes(y = y1, color = "p=0.3")) +
  geom_line(aes(y = y2, color = "p=3.4")) +
  geom_line(aes(y = y3, color = "p=8")) +
  labs(x = "x", y = "y") +
  theme(legend.position = c(0.8,0.1)) +
  scale_color_manual(name='p values',
                     breaks=c('p=0.3', 'p=3.4', 'p=8'),
                     values=c('p=0.3'='black', 'p=3.4'='red', 'p=8'='blue'))


exp5_plot
```

**Figure 9:** This plot shows how changes in parameter `p` affect the shape of exponential 5 curves, while holding other parameters constant.

