---
title: "Available Curves in tcplfit2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Available Curves in tcplfit2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(tcplfit2)
library(ggplot2)

data("mc3")
ex_conc <- seq(10^(min(mc3[,"logc"])),10^(max(mc3[,"logc"])),length.out = 100)
```

This vignette contains model details for all models available in `tcplfit2` package with parameter  explanations and plots.  

# Continuous Response Model

## Exponential 2
Model: $f(x) = a*(e^{\frac{x}{b}}-1) $

Parameters: 
1. a 
    + y-scale. If |a| > 1 and increases, the curve is stretched vertically. If 0 <|a| < 1 and decreases, the curve is shrunk vertically. Great summary from here. 
2. b
•	mathematical meaning: x-scale. If b >1 and increase, the curve is shrunk horizontally. If 0 < b < 1 and decrease, the curve is stretched horizontally. If b < 0, the curve is decreasing toward a horizontal asymptote x= -a.


```{r, exp2}
fits_exp2 <- data.frame(
  # base line 
  y1 = exp2(ps = c(a = 0.45,b = 12,er = 0.1), x = ex_conc),
  
  # change a
  y2 = exp2(ps = c(a = 1.5,b = 12,er = 0.1), x = ex_conc),
  y3 = exp2(ps = c(a = 0.1,b = 12,er = 0.1), x = ex_conc),
  
  # change b
  y4 = exp2(ps = c(a = 0.45,b = 13.5,er = 0.1), x = ex_conc),
  y5 = exp2(ps = c(a = 0.45,b = 9,er = 0.1), x = ex_conc)
  
)

exp2_plot1 <- ggplot(fits_exp2, aes(ex_conc)) +  
    geom_line(aes(y = y1, color = "a=0.45")) +
     geom_line(aes(y = y2, color = "a=0.6")) +
   geom_line(aes(y = y3, color = "a=0.1")) +
  labs(x = "x", y = "y") +
  theme(legend.position = c(0.1,0.8)) +
  scale_color_manual(name='a values',
                     breaks=c('a=0.45', 'a=0.6', 'a=0.1'),
                     values=c('a=0.45'='black', 'a=0.6'='red', 'a=0.1'='green'))

exp2_plot2 <- ggplot(fits_exp2, aes(ex_conc)) +  
    geom_line(aes(y = y1, color = "b=12")) +
     geom_line(aes(y = y4, color = "b=13.5")) +
   geom_line(aes(y = y5, color = "b=9")) +
  labs(x = "x", y = "y") +
  theme(legend.position = c(0.1,0.8)) +
  scale_color_manual(name='b values',
                     breaks=c('b=12', 'b=13.5', 'b=9'),
                     values=c('b=12'='black', 'b=13.5'='red', 'b=9'='green'))


exp2_plot1
exp2_plot2

```

## Exponential 3
Exp3 - a*(exp((x/b)^p - 1)
it's a and b are similar with those in Exponential 2, but it has one additional parameter p
1.	p 
•	mathematical meaning: power. A measure of how steep the curve is. 

```{r}
fits_exp3 <- data.frame(
  # base line 
  y1 = exp3(ps = c(a = 1.67,b = 12.5,p = 0.87,er = 0.1), x = ex_conc),
  
  # change a
  y2 = exp3(ps = c(a = 0.8,b = 12.5,p = 0.87,er = 0.1), x = ex_conc),
  y3 = exp3(ps = c(a = 2.5,b = 12.5,p = 0.87,er = 0.1), x = ex_conc),
  
  # change b
  y4 = exp3(ps = c(a = 1.67,b = 11,p = 0.87,er = 0.1), x = ex_conc),
  y5 = exp3(ps = c(a = 1.67,b = 14,p = 0.87,er = 0.1), x = ex_conc),
  
  # change p
  y6 = exp3(ps = c(a = 1.67,b = 12.5,p = 0.9,er = 0.1), x = ex_conc),
  y7 = exp3(ps = c(a = 1.67,b = 12.5,p = 0.7,er = 0.1), x = ex_conc)
  
)

exp3_plot1 <- ggplot(fits_exp3, aes(ex_conc)) +  
    geom_line(aes(y = y1, color = "a=1.67")) +
     geom_line(aes(y = y2, color = "a=0.8")) +
   geom_line(aes(y = y3, color = "a=2.5")) +
  labs(x = "x", y = "y") +
  theme(legend.position = c(0.1,0.8)) +
  scale_color_manual(name='a values',
                     breaks=c('a=1.67', 'a=0.8', 'b=9'),
                     values=c('a=1.67'='black', 'a=0.8' = 'red', 'a=2.5'='green'))

exp3_plot2 <- ggplot(fits_exp3, aes(ex_conc)) +  
    geom_line(aes(y = y1, color = "b=12.5")) +
     geom_line(aes(y = y4, color = "b=11")) +
   geom_line(aes(y = y5, color = "b=14")) +
  labs(x = "x", y = "y") +
  theme(legend.position = c(0.1,0.8)) +
  scale_color_manual(name='b values',
                     breaks=c('b=12.5', 'b=11', 'b=14'),
                     values=c('b=12.5'='black', 'b=11'='red', 'b=14'='green'))

exp3_plot3 <- ggplot(fits_exp3, aes(ex_conc)) +  
    geom_line(aes(y = y1, color = "p=0.87")) +
     geom_line(aes(y = y6, color = "p=0.9")) +
   geom_line(aes(y = y7, color = "p=0.7")) +
  labs(x = "x", y = "y") +
  theme(legend.position = c(0.1,0.8)) +
  scale_color_manual(name='p values',
                     breaks=c('p=0.87', 'p=0.9', 'p=0.7'),
                     values=c('p=0.87'='black', 'p=0.9'='red', 'p=0.7'='green'))

exp3_plot1
exp3_plot2
exp3_plot3
```

## Exponential 4
$y = \text{tp}*(1-2^{(\frac{-x}{\text{ga}})})$

1.	tp
•	Top. The horizontal asymptote the curve is approaching (also can be negative); the maximum or minimum of the predicted responses. 
2.	ga 
• AC50. Concentration at 50% of maximal activity, acts as slope for x, control how fast the curve is approaching the top. If ga increases, the curve is shrunk horizontally; if ga decrease, the curve is stretched horizontally. 


```{r}
fits_exp4 <- data.frame(
  # base line 
  y1 = exp4(ps = c(tp = 895,ga = 15,er = 0.1),x = ex_conc),
  
  # change tp
  y2 = exp4(ps = c(tp = 950,ga = 15,er = 0.1),x = ex_conc),
  y3 = exp4(ps = c(tp = 700,ga = 15,er = 0.1),x = ex_conc),
  
  # change ga
  y4 = exp4(ps = c(tp = 895,ga = 18,er = 0.1),x = ex_conc),
  y5 = exp4(ps = c(tp = 895,ga = 10,er = 0.1),x = ex_conc)
  
)

exp4_plot1 <- ggplot(fits_exp4, aes(ex_conc)) +  
    geom_line(aes(y = y1, color = "tp=895")) +
     geom_line(aes(y = y2, color = "tp=950")) +
   geom_line(aes(y = y3, color = "tp=700")) +
  labs(x = "x", y = "y") +
  theme(legend.position = c(0.1,0.8)) +
  scale_color_manual(name='tp values',
                     breaks=c('tp=895', 'tp=950', 'tp=700'),
                     values=c('tp=895'='black', 'tp=950'='red', 'tp=700'='green'))

exp4_plot2 <- ggplot(fits_exp4, aes(ex_conc)) +  
    geom_line(aes(y = y1, color = "ga=15")) +
     geom_line(aes(y = y4, color = "ga=18")) +
   geom_line(aes(y = y5, color = "ga=10")) +
  labs(x = "x", y = "y") +
  theme(legend.position = c(0.1,0.8)) +
  scale_color_manual(name='ga values',
                     breaks=c('ga=15', 'ga=18', 'ga=10'),
                     values=c('ga=15'='black', 'ga=18'='red', 'ga=10'='green'))


exp4_plot1
exp4_plot2
```


## exp5
Exp5 - tp*(1-2^(-(x/ga)^p))
tp and ga are similar to those in exp4, this has one additional 
1.	p 
•	mathematical meaning: power. A measure of how steep the curve is. 
 


```{r}
fits_exp5 <- data.frame(
  # base line 
  y1 = exp5(ps = c(tp = 793,ga = 6.25,p = 1.25,er = 0.1), x = ex_conc),
  
  # change tp
  y2 = exp5(ps = c(tp = 900,ga = 6.25,p = 1.25,er = 0.1), x = ex_conc),
  y3 = exp5(ps = c(tp = 500,ga = 6.25,p = 1.25,er = 0.1), x = ex_conc),
  
  # change ga
  y4 = exp5(ps = c(tp = 793,ga = 8.2,p = 1.25,er = 0.1), x = ex_conc),
  y5 = exp5(ps = c(tp = 793,ga = 6,p = 1.25,er = 0.1), x = ex_conc),
  
  # change p
  y6 = exp5(ps = c(tp = 793,ga = 6.25,p = 0.97,er = 0.1), x = ex_conc),
  y7 = exp5(ps = c(tp = 793,ga = 6.25,p = 1.5,er = 0.1), x = ex_conc)
  
)

exp5_plot1 <- ggplot(fits_exp5, aes(ex_conc)) +  
    geom_line(aes(y = y1, color = "tp=793")) +
     geom_line(aes(y = y2, color = "tp=900")) +
   geom_line(aes(y = y3, color = "tp=500")) +
  labs(x = "x", y = "y") +
  theme(legend.position = c(0.1,0.8)) +
  scale_color_manual(name='tp values',
                     breaks=c('tp=793', 'tp=900', 'tp=500'),
                     values=c('tp=793'='black', 'tp=900' = 'red', 'tp=500'='green'))

exp5_plot2 <- ggplot(fits_exp5, aes(ex_conc)) +  
    geom_line(aes(y = y1, color = "ga=6.25")) +
     geom_line(aes(y = y4, color = "ga=8.2")) +
   geom_line(aes(y = y5, color = "ga=6")) +
  labs(x = "x", y = "y") +
  theme(legend.position = c(0.1,0.8)) +
  scale_color_manual(name='ga values',
                     breaks=c('ga=6.25', 'ga=8.2', 'ga=6'),
                     values=c('ga=6.25'='black', 'ga=8.2'='red', 'ga=6'='green'))

exp5_plot3 <- ggplot(fits_exp5, aes(ex_conc)) +  
    geom_line(aes(y = y1, color = "p=1.25")) +
     geom_line(aes(y = y6, color = "p=0.99")) +
   geom_line(aes(y = y7, color = "p=1.5")) +
  labs(x = "x", y = "y") +
  theme(legend.position = c(0.1,0.8)) +
  scale_color_manual(name='p values',
                     breaks=c('p=1.25', 'p=0.97', 'p=1.5'),
                     values=c('p=1.25'='black', 'p=0.97'='red', 'p=1.5'='green'))

exp5_plot1
exp5_plot2
exp5_plot3
```


## polynomial 1 & 2

Polynomial 1 is linear line $y = ax$ where a is the slope of the line. 

polynimiao 2 in this package has slightly different parameterzation 

The normal polynimial 2 is $y = ax^2+bx+c$

The package is $a(\frac{x}{b} + \frac{x^2}{b^2})$. 

The vertex of a curve is determined by (-b/2, -a/4).
1.	a 
•	mathematical meaning: y scale. The sign of this parameter determines if the vertex will land below or below x-axis. If |a| >0 and increases, the curve is stretched vertically. If 0 <|a| < 1 and decreases, the curve is shrunk vertically.
2.	b 
•	mathematical meaning: x scale. The sign of this parameter determines if vertex will land on II/III or I/IV quadrants. B is restricted to be non-negative, so a's sign controls the directionality of the curve. If b increase, the curve is shrunk horizontally, and if b decrease, the curve is stretched horizontally. 

```{r}
fits_poly <- data.frame(
  # base line 
  y1 = poly1(ps = c(a = 3.5,er = 0.1),x = ex_conc),
  y1.2 = poly2(ps = c(a = 0.13,b = 2,er = 0.1),x = ex_conc),
  
  # change a for poly1
  y2 = poly1(ps = c(a = 5,er = 0.1),x = ex_conc),
  y3 = poly1(ps = c(a = 2.6,er = 0.1),x = ex_conc),

  # change a for poly2
  y4 = poly2(ps = c(a = 1, b = 2,er = 0.1),x = ex_conc),
  y5 = poly2(ps = c(a = 0.1, b = 2,er = 0.1),x = ex_conc),
  
  # change b for poly2
  y6 = poly2(ps = c(a = 0.13,b = 4,er = 0.1),x = ex_conc),
  y7 = poly2(ps = c(a = 0.13,b = 0.5,er = 0.1),x = ex_conc)
  
)

poly1_plot <- ggplot(fits_poly, aes(ex_conc)) +  
     geom_line(aes(y = y1, color = "a=3.5")) +
     geom_line(aes(y = y2, color = "a=5")) +
     geom_line(aes(y = y3, color = "a=2.6")) +
  labs(x = "x", y = "y") +
  theme(legend.position = c(0.1,0.8)) +
  scale_color_manual(name='a values',
                     breaks=c('a=3.5', 'a=5', 'a=2.6'),
                     values=c('a=3.5'='black', 'a=5' = 'red', 'a=2.6'='green'))

poly2_plot1 <- ggplot(fits_poly, aes(ex_conc)) +  
    geom_line(aes(y = y1.2, color = "a=0.13")) +
     geom_line(aes(y = y4, color = "a=1")) +
   geom_line(aes(y = y5, color = "a=0.1")) +
  labs(x = "x", y = "y") +
  theme(legend.position = c(0.1,0.8)) +
  scale_color_manual(name='a values',
                     breaks=c('a=0.13', 'a=1', 'a=0.1'),
                     values=c('a=0.13'='black', 'a=1'='red', 'a=0.1'='green'))

poly2_plot2 <- ggplot(fits_poly, aes(ex_conc)) +  
    geom_line(aes(y = y1.2, color = "b=2")) +
     geom_line(aes(y = y6, color = "b=4")) +
   geom_line(aes(y = y7, color = "b=0.5")) +
  labs(x = "x", y = "y") +
  theme(legend.position = c(0.1,0.8)) +
  scale_color_manual(name='b values',
                     breaks=c('b=2', 'b=4', 'b=0.5'),
                     values=c('b=2'='black', 'b=4'='red', 'b=0.5'='green'))

poly1_plot
poly2_plot1
poly2_plot2
```

## power 

$y=a*x^b$
1.	a
•	mathematical meaning: Scaling factor, moving the values of x^b up or down as it increases or decrease. Also control directionality. Also called slope in BMDS user guide.

2.	p
•	power. Rate of growth. Restricted to be larger than 0 or 1. 


```{r}
fits_pow <- data.frame(
  # base line 
  y1 = pow(ps = c(a = 1.23,p = 1.45,er = 0.1),x = ex_conc),
  
  # change a
  y2 = pow(ps = c(a = 3.2,p = 1.45,er = 0.1),x = ex_conc),
  y3 = pow(ps = c(a = 0.8,p = 1.45,er = 0.1),x = ex_conc),
  
  # change p
  y4 = pow(ps = c(a = 1.23,p = 1.5,er = 0.1),x = ex_conc),
  y5 = pow(ps = c(a = 1.23,p = 1.2,er = 0.1),x = ex_conc)
  
)

pow_plot1 <- ggplot(fits_pow, aes(ex_conc)) +  
     geom_line(aes(y = y1, color = "a=1.23")) +
     geom_line(aes(y = y2, color = "a=3.2")) +
     geom_line(aes(y = y3, color = "a=0.8")) +
  labs(x = "x", y = "y") +
  theme(legend.position = c(0.1,0.8)) +
  scale_color_manual(name='a values',
                     breaks=c('a=1.23', 'a=3.2', 'a=0.8'),
                     values=c('a=1.23'='black', 'a=3.2'='red', 'a=0.8'='green'))

pow_plot2 <- ggplot(fits_pow, aes(ex_conc)) +  
    geom_line(aes(y = y1, color = "p=1.45")) +
     geom_line(aes(y = y4, color = "p=1.5")) +
   geom_line(aes(y = y5, color = "p=1.2")) +
  labs(x = "x", y = "y") +
  theme(legend.position = c(0.1,0.8)) +
  scale_color_manual(name='p values',
                     breaks=c('p=1.45', 'p=1.5', 'p=1.2'),
                     values=c('p=1.45'='black', 'p=1.5'='red', 'p=1.2'='green'))


pow_plot1
pow_plot2
```


## Hill

Hill model is the only one has biological rational behind the parameter, included common interpretation of hill parameters here. 

1.	tp 
•	mathematical meaning: top, the horizontal asymptote the curve is approaching 
•	Biological meaning (if applicable): the highest or lowest response achieved at saturation
 
2.	ga 
•	mathematical meaning: AC50. Concentration at 50% of maximal activity
•	Biological meaning (if applicable): provides useful information about the "apparent affinity" of the protein under study (enzyme, transporter, etc.) for the substrate.
 
3.	P
•	mathematical meaning: Hill coefficient, acts as power, a measure of how steep the response curve is
•	Biological meaning (if applicable): a measure of the cooperativity of substrate binding to the enzyme, transporter, etc.


```{r}
fits_hill <- data.frame(
  # base line 
  y1 = hillfn(ps = c(tp = 750,ga = 5,p = 1.76,er = 0.1), x = ex_conc),
  
  # change tp
  y2 = hillfn(ps = c(tp = 850,ga = 5,p = 1.76,er = 0.1), x = ex_conc),

  # change ga
  y3 = hillfn(ps = c(tp = 750,ga = 7,p = 1.76,er = 0.1), x = ex_conc),
  y4 = hillfn(ps = c(tp = 750,ga = 4,p = 1.76,er = 0.1), x = ex_conc),
  
  # change p
  y5 = hillfn(ps = c(tp = 750,ga = 5,p = 2,er = 0.1), x = ex_conc),
  y6 = hillfn(ps = c(tp = 750,ga = 5,p = 1.5,er = 0.1), x = ex_conc)
  
)

hill_plot <- ggplot(fits_hill, aes(log10(ex_conc))) +  
    geom_line(aes(y = y1, color = "baseline")) +
     geom_line(aes(y = y2, color = "tp=850")) +
   geom_line(aes(y = y3, color = "ga=7")) +
  geom_line(aes(y = y4, color = "ga=4")) +
  geom_line(aes(y = y5, color = "p=2")) +
  geom_line(aes(y = y6, color = "p=1.5"))
  labs(x = "x", y = "y") +
  theme(legend.position = c(0.1,0.8)) +
  scale_color_manual(name='p values',
                     values=c('baseline'='black', 'tp=850'='red', 'ga=7'='green',
                              'ga=4' = 'blue', 'p=2'='pink', 'p=1.5'='yellow'))
  
hill_plot
```


## Gain-Loss

Gain-Loss is the product of two hill functions, one simulates the response going up, one simulates the responses going down. A curve can gain first and then loss or the opposite. 
1.	tp: maximum or minimum of the predicted response
2.	ga - gain AC50, the smaller the faster the curve approach to top (if increase first); 
3.	p - growth power, rate of gain
4.	la - loss AC50, the larger it gets the slower the curve starts to decrease (if it increase first)
5.	q - loss power, rate of loss 


```{r}
fits_gnls <- data.frame(
  # base line 
  y1 = gnls(ps = c(tp = 750,ga = 15,p = 1.45,la = 50,q = 1.34, er = 0.1), x = ex_conc),
  
  # change tp
  y2 = gnls(ps = c(tp = 750,ga = 15,p = 1.45,la = 50,q = 1.34, er = 0.1), x = ex_conc),

  # change ga
  y3 = gnls(ps = c(tp = 750,ga = 15,p = 1.45,la = 50,q = 1.34, er = 0.1), x = ex_conc),
  y4 = gnls(ps = c(tp = 750,ga = 15,p = 1.45,la = 50,q = 1.34, er = 0.1), x = ex_conc),
  
  # change p
  y5 = gnls(ps = c(tp = 750,ga = 15,p = 1.45,la = 50,q = 1.34, er = 0.1), x = ex_conc),
  y6 = gnls(ps = c(tp = 750,ga = 15,p = 1.45,la = 50,q = 1.34, er = 0.1), x = ex_conc)
  
)

gnls_plot <- ggplot(fits_gnls, aes(log10(ex_conc))) +  
    geom_line(aes(y = y1, color = "baseline")) +
     geom_line(aes(y = y2, color = "tp=850")) +
   geom_line(aes(y = y3, color = "ga=7")) +
  geom_line(aes(y = y4, color = "ga=4")) +
  geom_line(aes(y = y5, color = "p=2")) +
  geom_line(aes(y = y6, color = "p=1.5"))
  labs(x = "x", y = "y") +
  theme(legend.position = c(0.1,0.8)) +
  scale_color_manual(name='p values',
                     values=c('baseline'='black', 'tp=850'='red', 'ga=7'='green',
                              'ga=4' = 'blue', 'p=2'='pink', 'p=1.5'='yellow'))
  
gnls_plot
```



