---
title: "tcplfit2 Vignette"
author: "Center for Computational Toxicology and Exposure"
output:
   prettydoc::html_pretty:
    theme: architect
    toc: yes
    toc_depth: 4
vignette: >
  %\VignetteIndexEntry{tcplfit2_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{css, echo=FALSE}
.scroll-100 {
  max-height: 100px;
  overflow-y: auto;
}
```

```{r setup, warning = FALSE, echo = FALSE}
library(tcplfit2)
```
# Getting started with tcplfit2
The package tcplfit2 contains the core concentration-response functionality of the package tcpl (The ToxCast Pipeline) built to process all of the ToxCast high-throughput screen (HTS) data at the US EPA. Much of the rest of the code in tcpl is used to do data processing, normalization, and database storage. We wanted to reuse the core concentration-response code for other projects, and add extensions to it, which was the origin of the current package `tcplfit2`. The main set of extensions was to include all of the concentration-response models that are contained in the program [BMDExpress](https://www.sciome.com/bmdexpress/). 
These include exponential, polynomial (1 & 2), and power functions in addition to the original Hill, gain-loss and constant models.
Additionally, we wanted to include BMD (Benchmark Dose Modeling) outputs, which is simply defining a 
Benchmark Response (BMR) level and setting the BMD to the concentration where the curve crosses the BMR level. 
One final addition was to let the hitcall value be a continuous number ranging from 0 to 1. Continuous hitcall in `tcplfit2` are defined as the product of three proportional weights: 1) the AIC of the winning model is better than the constant model (i.e. winning model is not fit to background noise), 2) at least one concentration has a median response that exceeds cutoff, and 3) the top from the winning model exceeds the cutoff. This vignette describes some functionality of the `tcplfit2` package with a few simple examples.

## Example 1: Running a single concentration-response calculation
All calculations use the function `concRespCore` which has several key inputs. The first set are 
put into a named list called 'row':

* `conc` - a vector of concentrations (not log concentrations)
* `resp` - a vector of responses, of the same length as `conc`. Note that replicates are allowed, i.e. there can be multiple pairs of conc and resp with the same concentration value.
* `cutoff`- this is the value that the response must exceed before a a curve can be called a hit. For ToxCast, this is usually some multiple (typically 3) of the median absolute deviation (BMAD) around baseline for the lowest two concentration. The user is free to make other choices
* `bmed` - this is the median of the baseline. The entire response series will be shifted by this amount. Set to zero if the data is zero-centered.  
* `onesd`- This is one standard deviation of the noise around the baseline. The BMR value = `onesd`*`bmr_scale`. The default `bmr_scale` is 1.349.

The function `concRespCore` can also have other optional elements which will be included in the output. These can be,
for instance, the name of the chemical (or other identifiers) or the name of the assay being modeled. Two other 
parameters might be used. The first is a Boolean `conthits`. If TRUE (the default, and recommended usage), the hitcall
returned will be a continuous value between 0 and 1. The other is `do.plot`. If this is set to TRUE (default is FALSE),
a plot of the curve will be generated. The user can also select only a subset of the models to be run. The example below has all of the possible ones
included. If the `fitmodels` parameter is specified, then it must always include the constant model (`cnst`) since this provides one basis for comparison during hitcalling.  However, some models may be excluded, for example the gain-loss (`gnls`) model is excluded in some applications.

To run a simple example, use the following code ...
```{r example1, fig.height = 4.55, fig.width = 8}
  conc <- list(.03,.1,.3,1,3,10,30,100)
  resp <- list(0,.2,.1,.4,.7,.9,.6, 1.2)
  row = list(conc = conc, resp = resp, bmed = 0, cutoff = 1, onesd = .5,name="some chemical")
  oldpar <- par(no.readonly = TRUE)
  on.exit(par(oldpar))   
  par(xpd = TRUE)
  res <- concRespCore(row,fitmodels = c("cnst", "hill", "gnls", "poly1", "poly2", "pow", "exp2", "exp3",
                                        "exp4", "exp5"),conthits = T, do.plot=T)
  
```

The output of this run will be a data frame with one row, summarizing the results for the winning model.

```{r example1 result, warning=FALSE, echo=FALSE}
library(DT)
DT::datatable(res,rownames = FALSE,options = list(scrollX = T))
```

## Example 2: Running a series of concentration-response models for a single assay
The input data for this example is taken from one of the Tox21 HTS assays, for estrogen receptor (ER) 
agonist activity. The data is from the mc3 table in the database `invitrodb`, which is the back end for
`tcpl`. The input data for this example have already been through the pre-processing steps (prior to `tcpl`), and processed through several steps in the ToxCast pipeline (including normalization of concentration data and transformation of response values - Levels 0 - 3). This example will run 6 chemicals out of the 100 that are included in the data set, and will create plots 
for these. The plotting routine `concRespPlot` is somewhat generic, and we anticipate that users will make
their own version of this. To run this example, use the following code ...
```{r example2, fig.height = 8, fig.width = 7}
  # read in the data
  # Loading in the level 3 example data set from invitrodb
  data("mc3")

  # set up a 3 x 2 grid for the plots
  oldpar <- par(no.readonly = TRUE)
  on.exit(par(oldpar))            
  par(mfrow=c(3,2),mar=c(4,4,2,2))

  # determine the background variation
  temp <- mc3[mc3$logc<= -2,"resp"]
  bmad <- mad(temp)
  onesd <- sd(temp)
  cutoff <- 3*bmad

  # select six samples. Note that there may be more than one sample processed for a given chemical
  spid.list <- unique(mc3$spid)
  spid.list <- spid.list[1:6]

  for(spid in spid.list) {
    # select the data for just this sample
    temp <- mc3[is.element(mc3$spid,spid),]

    # The data file has stored concentration in log10 form, so fix that
    conc <- 10**temp$logc
    resp <- temp$resp

    # pull out all of the chemical identifiers and the name of the assay
    dtxsid <- temp[1,"dtxsid"]
    casrn <- temp[1,"casrn"]
    name <- temp[1,"name"]
    assay <- temp[1,"assay"]

    # create the row object
    row <- list(conc = conc, resp = resp, bmed = 0, cutoff = cutoff, onesd = onesd,assay=assay,dtxsid=dtxsid,casrn=casrn,name=name)

    # run the concentration-response modeling for a single sample
    res <- concRespCore(row,fitmodels = c("cnst", "hill", "gnls", "poly1", "poly2", "pow", "exp2", "exp3",
                                          "exp4", "exp5"),conthits = T, aicc = F,bidirectional=F)

    # plot the results
    concRespPlot(res,ymin=-10,ymax=100)
  }


```

One would typically save the result rows in a data frame end export these for further analysis. You could remove the plotting function from the current loop and have a loop that read from the overall results data frame and only plot selected results (e.g. those with significant responses).

```{r example2 result, warning=FALSE, echo=FALSE}
DT::datatable(res,rownames = FALSE,options = list(scrollX = T))
```

## Example 3: Plotting concentration-response modeling on transcriptional signatures
The input data for this example contains 6 signatures for one chemical in a transcriptomics data set. Each signature is a different assay endpoint, thus one row in the data represents a given chemical and signature pair (assay endpoint). This data set is a sample from the signature scoring method that provides the cutoff, one standard deviation, and the concentration-response data. The example illustrates two kinds of plots available in `tcplfit2`. In the call to `concRespCore()`, the argument `do.plot` is set to `TRUE`, which provides a simple plot showing results of all the different curve fitting methods. Next, utilizing the function `concRespPlot()` provides a more informative plot for the winning model.
```{r example3, fig.height = 6, fig.width = 7, warning = FALSE}
  # call additional R packages
  library(stringr)  # string management package

  # read in the file
  data("signatures")
  
  # set up a 3 x 2 grid for the plots
  oldpar <- par(no.readonly = TRUE)
  on.exit(par(oldpar))            
  par(mfrow=c(3,2),mar=c(4,4,2,2))
    
  # fit 6 observations in signatures
  for(i in 1:nrow(signatures)){
    # set up input data
    row = list(conc=as.numeric(str_split(signatures[i,"conc"],"\\|")[[1]]),
               resp=as.numeric(str_split(signatures[i,"resp"],"\\|")[[1]]),
               bmed=0,
               cutoff=signatures[i,"cutoff"],
               onesd=signatures[i,"onesd"],
               name=signatures[i,"name"],
               assay=signatures[i,"signature"])
    # run concentration-response modeling (1st plotting option)
    out = concRespCore(row,conthits=F,do.plot=T)
    if(i==1){
      res <- out
    }else{
      res <- rbind.data.frame(res,out)
    }
  }
```

```{r example3_plot2, fig.height = 8, fig.width = 7}
  # set up a 3 x 2 grid for the plots
  oldpar <- par(no.readonly = TRUE)
  on.exit(par(oldpar))            
  par(mfrow=c(3,2),mar=c(4,4,2,2))
  # plot results using `concRespPlot`(2nd plotting option)
  for(i in 1:nrow(res)){
    concRespPlot(res[i,],ymin=-1,ymax=1)
  }
```

## Example 4: Running tcpl-like multi-concentration response data without a database connection
The ToxCast pipeline `tcpl` is an R package that manages, curve-fits, plots, and stores ToxCast data to populate its linked MySQL database, InvitroDB. The original `tcplFit()` function within `tcpl` performed basic concentration response curve fitting. Processing with tcpl_v3 and beyond depends on `tcplfit2` to allow a wider variety of concentration-response models when using `invitrodb` in the 4.0 schema and beyond. `tcplLite` was deprecated with the updates to `tcpl` and development of `tcplfit2`, since `tcplfit2` allows one to perform curve-fitting and hit-calling independent of a database. The example below demonstrates how to perform an analogous `tcplLite` analysis with `tcplfit2`. For additional information, please consult vignettes for `library(tcpl)` at https://CRAN.R-project.org/package=tcpl. 

The input for this example comes from the ACEA_AR assay. Data from the assay component ACEA_AR_agonist_80hr was analyzed in the positive analysis fitting direction relative to DMSO as the neutral control and baseline of activity. Using a electrical impedance as a cell growth reporter, increased activity can be used to infer increased signaling at the pathway-level for the androgen receptor (as encoded by the AR gene). Given heterogeneous assay data, source data often must go through pre-processing steps to transform into a uniform data format, often like this level 0. The below table is identical to the multi-concentration level 0 data (mc0) table that would be seen in `invitrodb` and recognized by `tcpl`. Columns include:

* m0id = Level 0 id
* spid = Sample id
* acid = Unique assay component id; unique numeric id for each assay component
* apid = Assay plate id
* coli = Column index (location on assay plate)
* rowi = Row index (location on assay plate)
* wllt = well type
* wllq = well quality
* conc = concentration
* rval = raw value
* srcf = Source file name
* clowder_uid = clowder unique id for source files
* git_hash = hash key for pre-processing scripts

```{r example4_init, fig.height = 6, fig.width = 7, message=FALSE, warning = FALSE}
# Loading in the level 0 example data set from invitrodb
data("mc0")
library(data.table)
dat <- mc0
DT::datatable(head(dat[wllt=='t',]),rownames= FALSE, options = list(scrollX = T))
```

To run standalone `tcplfit2` fitting without the need for a MySQL database connection like `invitrodb`, the user will replicate stepping through the multiple levels of processing. A detailed explanation of processing levels can be found within `tcpl`'s Data Processing vignette.

Level 1 importantly establishes the concentration index. The concentration index is simply the distinct concentrations ranked from lowest to highest, and this index can be used to calculate the baseline median absolute deviation for an assay.

```{r example4_cndx, fig.height = 6, fig.width = 7}
library(tcpl)
## Order by the following columns
setkeyv(dat, c('acid', 'srcf', 'apid', 'coli', 'rowi', 'spid', 'conc'))

## Define replicate id (rpid) column for test compound wells
nconc <- dat[wllt == "t" , ## denotes test well as the well type (wllt)
             list(n = lu(conc)), #total number of unique concentrations
             by = list(acid, apid, spid)][ , list(nconc = min(n)), by = acid]
dat[wllt == "t" & acid %in% nconc[nconc > 1, acid],
    rpid := paste(acid, spid, wllt, srcf, apid, "rep1", conc, sep = "_")]
dat[wllt == "t" & acid %in% nconc[nconc == 1, acid],
    rpid := paste(acid, spid, wllt, srcf, "rep1", conc, sep = "_")]

## Define rpid column for non-test compound wells
dat[wllt != "t",
    rpid := paste(acid, spid, wllt, srcf, apid, "rep1", conc, sep = "_")]

## set repid based on rowid
dat[, dat_rpid := rowid(rpid)]
dat[, rpid := sub("_rep[0-9]+.*", "",rpid, useBytes = TRUE)]
dat[, rpid := paste0(rpid,"_rep",dat_rpid)]

# Define concentration index
indexfunc <- function(x) as.integer(rank(unique(x))[match(x, unique(x))])
dat[ , cndx := indexfunc(conc), by = list(rpid)]
```

### Adjustments
Levels 2 and 3 are used for data adjustments and normalization. Generally if the response values (`rval`) need to be logged or transformed in some way from their original values this is where that adjustment would occur. Transformed response values are referred to as corrected values and are stored in the `cval` field/variable. However, in this case, the corrected values (`cval`) are identical to the original response values (`rval`). 

```{r example4_mc2, fig.height = 6, fig.width = 7}
# If no adjustments are required for the data, the corrected value (cval) should be set as original rval
dat[,cval := rval]

## Poor well quality (wllq) wells should be removed
dat <- dat[!wllq == 0,]

## Fitting generally cannot occur if response values are NA therefore values need to be removed
dat <- dat[!is.na(cval),]

## A column for log10 concentration is added as some of the mc3 methods require logc. Given logging concentration, conc=0 are not allowed therefore a dummy non-zero value should be used
dat[conc == 0 , conc := 0.0001]
dat[ , logc := log10(conc)]

#As a final step to prepare the dataset tcplfit2 processing, a dummy aeid is required if using mc3_mthds from tcpl
dummy_aeid <- 99999
dat[,aeid := dummy_aeid]

## Set aeid as a key
setkey(dat,aeid)
```

Once the data is initialized to a point where the required fields are available, the methods included in the `tcpl` package can be identified and applied without the need for a database connection. You can see the list of available methods for Level 3 in the table below:

```{r example4_mthdlist, fig.height = 6, fig.width = 7, warning = FALSE}
mthd_funcs <- tcpl:::mc3_mthds()
DT::datatable(tcpl::tcplMthdList(3),rownames= FALSE, options = list(scrollX = T))
```

### Normalization
Here three normalization methods are selected and applied to the data.  Note because of the way `tcpl` handles the application of functions, the dataframe must be called `dat`. In the future, `tcpl` will export these functions so that they can be applied to any dataset without the need for a specific name or dummy aeid.

```{r example4_mc3methods, fig.height = 6, fig.width = 7, results = 'hide'}
# apply level 3 methods
## These methods directly apply the normalization methods from tcpl without the need for a DB connection
lapply(mthd_funcs[["bval.apid.nwlls.med"]](dummy_aeid), eval)
lapply(mthd_funcs[["pval.apid.medncbyconc.min"]](dummy_aeid),eval)
lapply(mthd_funcs[["resp.pc"]](dummy_aeid),eval)
```

Level 4 determines the baseline variability, or noise, that will later be used for cutoff calculation. Using the established concentration index, the level 4 methods can be loaded in a similar way to level 3.

```{r example4_mthdlist_l4, fig.height = 6, fig.width = 7}
mthd_funcs_l4 <- tcpl:::mc4_mthds()
DT::datatable(tcpl::tcplMthdList(4), rownames= FALSE, options = list(scrollX = T))
```

There are much fewer level 4 methods, but generally it is a requirement to assign a method that calculates the bmad and assign a method that calculates the standard deviation of the noise for `tcplfit2` fitting.

```{r example4_mc4methods, fig.height = 6, fig.width = 7, results = 'hide'}
# apply level 4 methods
## These methods directly apply the noise calculation and fitting methods from tcpl without the need for a DB connection
lapply(mthd_funcs_l4[["bmad.aeid.lowconc.twells"]](),eval)
lapply(mthd_funcs_l4[["onesd.aeid.lowconc.twells"]](),eval)
lapply(mthd_funcs_l4[["bidirectional.false"]](),eval)
```

### Dose-Response Curve Fitting
After methods up to level 4 have been applied, the model fitting can begin. In `tcpl`, this would be considered level 4, and is where `tcplfit2` is used to fit all of the models as a dependency for `tcpl`.

```{r example4_fitting, fig.height = 6, fig.width = 7}
#do tcplfit2 fitting
myfun <- function(y) {
  res <- tcplfit2::tcplfit2_core(y$conc,
                          y$resp,
                          cutoff = unique(y$bmad),
                          bidirectional = TRUE,
                          verbose = FALSE,
                          force.fit = TRUE,
                          fitmodels = c("cnst", "hill", "gnls", "poly1",
                                        "poly2", "pow", "exp2", "exp3",
                                        "exp4", "exp5")
                          )
  list(list(res)) #use list twice because data.table uses list(.) to look for values to assign to columns
}
```

The following code performs dose-response modeling for all spids in the dataset. **Warning: The fitting step for the full data set, `dat`, can take 7-10 minutes to run.** Hence the code chunk following provides a subset example of data for curve fitting and hitcalling. The subset data only contains records of six samples.

```{r example4_fitting_full, eval=FALSE}
# only want to run tcplfit2 for test wells in this case
# this chunk doesn't run, fit the curves on the subset below
dat[wllt == 't',params:= myfun(.SD), by = .(spid)]
```

```{r example4_fitting_subset}
# create a subset that contains 6 samples and run curve fitting
subdat <- dat[spid %in% unique(spid)[10:15],]
subdat[wllt == 't',params:= myfun(.SD), by = .(spid)]
```

### Continuous Hitcalling
After all of the models have been fit, hitcalling can occur. The output of level 4 can be fed directly into the `tcplhit2_core` function. The results are then pivoted and shown in the resulting datatable.

```{r example4_hitcalling, fig.height = 6, fig.width = 7}
myfun2 <- function(y) {
  res <- tcplfit2::tcplhit2_core(params = y$params[[1]],
                                 conc = y$conc,
                                 resp = y$resp,
                                 cutoff = 3*unique(y$bmad),
                                 onesd = unique(y$osd)
                                 )
  list(list(res))
}

# continute with hitcalling
res <- subdat[wllt == 't', myfun2(.SD), by = .(spid)]

#pivot wider
res_wide <- rbindlist(Map(cbind, spid = res$spid, res$V1))

DT::datatable(res_wide,options = list(scrollX = T))
```

*Hitcalling can also be done with the full data set, `dat`.*

The output table resulting from the previous code chunk is the same format as the `res` table in example 3. Thus, one can use the `concRespPlot` function, as done previously in example 3, to plot the results. The next code chunk demonstrates how to visualize the example 4 fit results.

```{r example4_plot, fig.height = 8, fig.width = 7}
  # set up a 3 x 2 grid for the plots
  oldpar <- par(no.readonly = TRUE)
  on.exit(par(oldpar))            
  par(mfrow=c(3,2),mar=c(4,4,2,2))
  # plot results using `concRespPlot`(2nd plotting option)
  for(i in 1:nrow(res)){
    concRespPlot(res_wide[i,],ymin=-50,ymax=50)
  }
```

## Example 5: How to "Censor" BMD
The estimated BMD can occur above or below an experimental dose or other specified values; e.g., it is higher or lower than the maximum or minimum tested concentrations in the data. When this happens, `tcplhit2_core` and `concRespCore` provide options for users to "censor" the estimated BMD with the use of arguments `bmd_low_bnd` and `bmd_up_bnd`. 

`bmd_low_bnd` and `bmd_up_bnd` are multipliers applied to the minimum or maximum tested concentrations (i.e. reference doses) to generate the lower and upper boundaries for BMD estimates (i.e. threshold doses), respectively. This section provides examples to demonstrate how to "censor" BMD estimates with a lower and an upper bound, and how to "censor" BMD estimates with calls to `concRespCore` and `tcplhit2_core`.

### "Censor" BMD with a lower bound {#censorlowerbound}
First, let's consider cases when the estimated BMD is below the lowest tested concentration. When this happened, it implies that the experiment was not designed to have enough low-dose groups to capture behaviors in the low-dose regions and it can be a red flag for unreliable analysis. 

In the following code chunk, we use some example data from `mc3` to simulate an example of this case. We take the data of a sample and remove data from dose groups between 0 and 0.41, the 

remianing data set contains data for dose groups of 0.41 or higher, and the remianing reponses shows large increases in the fitst couples of dose groups. 
```{r example 5 lower}
# We'll use data from mc3 in this section
data("mc3")

# determine the background variation
temp <- mc3[mc3$logc<= -2,"resp"]
bmad <- mad(temp)
onesd <- sd(temp)
cutoff <- 3*bmad

# load example data
spid <- unique(mc3$spid)[94]
ex_df <- mc3[is.element(mc3$spid,spid),]

# The data file has stored concentration in log10 form, fix it 
conc <- 10**ex_df$logc # back-transforming concentrations on log10 scale
resp <- ex_df$resp

# modify the data for demonstration purposes 
conc2 <- conc[conc>0.41]
resp2 <- resp[which(conc>0.41)]

# pull out all of the chemical identifiers and the name of the assay
dtxsid <- ex_df[1,"dtxsid"]
casrn <- ex_df[1,"casrn"]
name <- ex_df[1,"name"]
assay <- ex_df[1,"assay"]

# create the row object
row_low <- list(conc = conc2, resp = resp2, bmed = 0, cutoff = cutoff, onesd = onesd,
            assay=assay, dtxsid=dtxsid,casrn=casrn,name=name)

# run the concentration-response modeling for a single sample
res_low <- concRespCore(row_low,fitmodels = c("cnst", "hill", "gnls", "poly1", "poly2", 
                                          "pow", "exp2", "exp3", "exp4", "exp5"), 
                        conthits = T, aicc = F, bidirectional=F)

concRespPlot(res_low,ymin=-10,ymax=60)
```

```{r example 5 lower-res}
# function results
res_low['Min. Conc.'] <- min(conc2)
res_low['Name'] <- name
res_low[1, c("Min. Conc.", "bmd", "bmdl", "bmdu")] <- round(res_low[1, c("Min. Conc.", "bmd", "bmdl", "bmdu")], 3)
DT::datatable(res_low[1, c("Name","Min. Conc.", "bmd", "bmdl", "bmdu")],rownames = FALSE)
```

The lowest concentration (i.e. reference dose) in the data is 0.6 but the estimated BMD from the hit-calling result is 0.302, which is lower. Users can allow the estimated BMD to be lower than the reference dose while restricting it to be no lower than a threshold dose set by using the argument `bmd_low_bnd`. 

For example, we want the estimated BMD to be no lower than 80% of the reference dose. We can use `bmd_low_bnd = 0.8` to set a threshold dose of 0.48. If the estimated BMD is lower than the threshold dose (like in this example), it will be "censored" to the threshold dose, and its confident boundaries will be shifted right by a distance equals to the difference between the estimated BMD and the threshold dose. Figure 7 provides a visual representation of this process. Valid inputs for `bmd_low_bnd` are numbers between 0 and 1, not including 0.
```{r example 5 lower-demo}
# using the argument to set a lower bound for BMD
res_low2 <- concRespCore(row_low,fitmodels = c("cnst", "hill", "gnls", "poly1", "poly2", 
                                           "pow", "exp2", "exp3", "exp4", "exp5"), 
                         conthits = T, aicc = F, bidirectional=F, bmd_low_bnd = 0.8)
```

```{r example 5 new lower-res}
# print out the new results
# include previous results side by side for comparison 
res_low2['Min. Conc.'] <- min(conc2)
res_low2['Name'] <- paste(name, "after `censoring`", sep = "-")
res_low['Name'] <- paste(name, "before `censoring`", sep = "-")
res_low2[1, c("Min. Conc.", "bmd", "bmdl", "bmdu")] <- round(res_low2[1, c("Min. Conc.", "bmd", "bmdl", "bmdu")], 3)

output_low <- rbind(res_low[1, c('Name', "Min. Conc.", "bmd", "bmdl", "bmdu")], 
                    res_low2[1, c('Name', "Min. Conc.", "bmd", "bmdl", "bmdu")])
DT::datatable(output_low,rownames = FALSE)
```

```{r example 5 lower plot, class.source="scroll-100"}
# generate some concentration for the fitted curve 
logc_plot <- seq(from=-3,to=2,by=0.05)
conc_plot <- 10**logc_plot

# initiate the plot
plot(conc2,resp2,xlab="conc (uM)",ylab="Response",xlim=c(0.001,100),ylim=c(-5,60),
       log="x",main=paste(name,"\n",assay),cex.main=0.9)

# add vertical lines to mark the minimum concentration in the data and the lower threshold set by bmd_low_bnd
abline(v=min(conc2), lty = 1, col = "brown", lwd = 2)
abline(v=res_low2$bmd, lty = 2, col = "darkviolet", lwd = 2)

# add markers for BMD and its boundaries before `censoring`
lines(c(res_low$bmd,res_low$bmd),c(0,50),col="green",lwd=2)
rect(xleft=res_low$bmdl,ybottom=0,xright=res_low$bmdu,ytop=50,col=rgb(0,1,0, alpha = .5), border = NA)
points(res_low$bmd, 0, pch = "x", col = "green")

# add markers for BMD and its boundaries after `censoring`
lines(c(res_low2$bmd,res_low2$bmd),c(0,50),col="blue",lwd=2)
rect(xleft=res_low2$bmdl,ybottom=0,xright=res_low2$bmdu,ytop=50,col=rgb(0,0,1, alpha = .5), border = NA)
points(res_low2$bmd, 0, pch = "x", col = "blue")

# add the fitted curve
lines(conc_plot, exp4(ps = c(res_low$tp, res_low$ga), conc_plot))
legend(1e-3, 60, legend=c("Reference Dose", "Threshold Dose", "BMD-before", "BMD-after"),
       col=c("brown", "darkviolet", "green", "blue"), lty=c(1,2,1,1))
```

**Figure 7**: This plot shows how BMD and its boundaries changed after "censoring". The green line and green "X" mark the estimated BMD before "censoring" and the green column represents its boundaries. The blue line and blue "X" mark the estimated BMD after "censoring" and the blue column represents its boundaries. The brown solid line represents the minimum concentration in the data, and the dark dashed line represents the threshold dose set by `bmd_low_bnd`. In this example, BMD was "censored" to the threshold value, and its boundaries were shifted right.

### "Censor" BMD with an upper bound 
Let's consider another situation when BMD is higher than the maximum concentration in the data. This is not ideal because the experiments are typically designed as that anything beyond the maximum concentrate would cause server damage. If this happen, `bmd_up_bnd` can be used to restrict the BMD to be no higher than a percentage of the maximum dose. 
```{r example 5 upper}
# load example data
spid <- unique(mc3$spid)[26]
ex_df <- mc3[is.element(mc3$spid,spid),]

# The data file has stored concentration in log10 form, so fix that
conc <- 10**ex_df$logc # back-transforming concentrations on log10 scale
resp <- ex_df$resp

# pull out all of the chemical identifiers and the name of the assay
dtxsid <- ex_df[1,"dtxsid"]
casrn <- ex_df[1,"casrn"]
name <- ex_df[1,"name"]
assay <- ex_df[1,"assay"]

# create the row object
row_up <- list(conc = conc, resp = resp, bmed = 0, cutoff = cutoff, onesd = onesd,assay=assay,
            dtxsid=dtxsid,casrn=casrn,name=name)

# run the concentration-response modeling for a single sample
res_up <- concRespCore(row_up,fitmodels = c("cnst", "hill", "gnls", "poly1", "poly2", 
                                         "pow", "exp2", "exp3", "exp4", "exp5"), 
                       conthits = T, aicc = F, bidirectional=F)

concRespPlot(res_up,ymin=-10,ymax=40)
```

```{r example 5 upper-res}
# max conc.
res_up['Max Conc.'] <- max(conc)
res_up['Name'] <- name
res_up[1, c("Max Conc.", "bmd", "bmdl", "bmdu")] <- round(res_up[1, c("Max Conc.", "bmd", "bmdl", "bmdu")], 3)
# function results
DT::datatable(res_up[1, c('Name','Max Conc.', "bmd", "bmdl", "bmdu")],rownames = FALSE)
```

The BMD 299.927 is higher than the maximum concentration in the data, which is 80. One can set `bmd_up_bnd` to 1 to use the maximum concentration as the upper bound (i.e., BMD is not allowed to be larger than 100% of the maximum concentration in the data). And if BMD is bigger than this upper bound, it will be forced to be the maximum concentration. One can also input a number larger than 1 to `bmd_up_bnd` so BMD is allowed to be larger than the maximum concentration but not larger than a specified number. For example, set it to 2 does not allow BMD to be larger than 2 times the maximum concentration. Doing this can force the extreme BMD to be topped by a specified value. 
```{r example upper-demo}
# using bmd_up_bnd = 2
res_up2 <- concRespCore(row_up,fitmodels = c("cnst", "hill", "gnls", "poly1", "poly2", 
                                          "pow", "exp2", "exp3", "exp4", "exp5"), 
                        conthits = T, aicc = F, bidirectional=F, bmd_up_bnd = 2)
```

```{r example upper-2}
# print out the new results
# include previous results side by side for comparison 
res_up2['Max Conc.'] <- max(conc)
res_up2['Name'] <- paste(name, "after `censoring`", sep = "-")
res_up['Name'] <- paste(name, "before `censoring`", sep = "-")
res_up2[1, c("Max Conc.", "bmd", "bmdl", "bmdu")] <- round(res_up2[1, c("Max Conc.", "bmd", "bmdl", "bmdu")], 3)

output_up <- rbind(res_up[1, c('Name', "Max Conc.", "bmd", "bmdl", "bmdu")], 
                   res_up2[1, c('Name', "Max Conc.", "bmd", "bmdl", "bmdu")])
DT::datatable(output_up,rownames = FALSE)
```

```{r example upper plot, class.source="scroll-100"}
# generate some concentration for the fitting curve 
logc_plot <- seq(from=-3,to=2,by=0.05)
conc_plot <- 10**logc_plot

# initiate plot
plot(conc,resp,xlab="conc (uM)",ylab="Response",xlim=c(0.001,500),ylim=c(-5,60),
       log="x",main=paste(name,"\n",assay),cex.main=0.9)
# add vertical lines to mark the maximum concentration in the data and the upper threshold set by bmd_up_bnd
abline(v=max(conc), lty = 2, col = "black", lwd=2)
abline(v=160, lty = 2, col = "darkviolet", lwd=2)

# add marker for BMD and its boundaries before `censoring`
lines(c(res_up$bmd,res_up$bmd),c(0,50),col="green",lwd=2)
rect(xleft=res_up$bmdl,ybottom=0,xright=res_up$bmdu,ytop=50,col=rgb(0,1,0, alpha = .5), border = NA)
points(res_up$bmd, 0, pch = "x", col = "green")

# add marker for BMD and its boundaries after `censoring`
lines(c(res_up2$bmd,res_up2$bmd),c(0,50),col="blue",lwd=2)
rect(xleft=res_up2$bmdl,ybottom=0,xright=res_up2$bmdu,ytop=50,col=rgb(0,0,1, alpha = .5), border = NA)
points(res_up2$bmd, 0, pch = "x", col = "blue")

# add the fitting curve
lines(conc_plot, poly1(ps = c(res_up$a), conc_plot))
```

**Figure 9**: This plot shows how BMD and its boundaries changed after "censoring". The green line and green "X" mark the BMD before "censoring" and the green column shows the original boundaries. The blue line and blue "X" mark the "censored" BMD and the blue column shows the updated boundaries. The brown dash line represents the maximum concentration in the data, and the violet dash line represents the threshold value set by `bmd_up_bnd`. In this example, BMD was "censored" to the threshold value, and its boundaries were shifted left.

### "Censor" BMD with tcplhit2_core
`bmd_low_bnd` and `bmd_low_bnd` are arguments present in `tcplhit2_core`. `concRespCore` "inherit" them because it leverages `tcplhit2_core` within the function. That means, if one chooses to use `tcplfit2_core` and `tcplhit2_core` separately, they can do the same "censoring" for BMD using `tcplhit2_core`. The results from these two methods are identical. Some example codes are provided below to show how to get the same result from the lower bound example above with `tcplhit2_core`.
```{r example with hit core}
# using the same data, fit curves 
param <- tcplfit2_core(conc2, resp2, cutoff = cutoff)
hit_res <- tcplhit2_core(param, conc2, resp2, cutoff = cutoff, onesd = onesd, 
                         bmd_low_bnd = 0.8)
```

```{r res-hit core}
# adding the result from tcplhit2_core to the output table for comparison
hit_res["Name"]<-  paste("Chlorothalonil", "tcplhit2_core", sep = "-")
hit_res['Min. Conc.'] <- min(conc2)
hit_res[1, c("Min. Conc.", "bmd", "bmdl", "bmdu")] <- round(hit_res[1, c("Min. Conc.", "bmd", "bmdl", "bmdu")], 3)

output_low <- rbind(output_low, 
                    hit_res[1, c('Name', "Min. Conc.", "bmd", "bmdl", "bmdu")])
DT::datatable(output_low,rownames = FALSE)
```

### BMD falls between the "threshold value" and the "reference dose"
If the estimated BMD falls between the minimum or maximum concentration and the threshold bounds set by `bmd_low_bnd` or `bmd_up_bnd`, the result will be returned as-is. For demonstration purposes we use the lower bound censoring example, but the same principle applies for the upper bound censoring case.

Codes below take the example from section ["censor" with lower bound](#censorlowerbound) and show what happens if we use a lower threshold bound. Let's assume it is acceptable as long as the estimated BMD is not lower than 40% (two fifths) of the minimum concentration in the data. The estimated BMD 0.302 is lower than the minimum concentration 0.6, but is higher than 40% of 0.6, which is 0.24. In this case, the BMD estimates and its boundaries will be changed. 
```{r example even lower bound}
res_low3 <- concRespCore(row_low,fitmodels = c("cnst", "hill", "gnls", "poly1", "poly2", 
                                           "pow", "exp2", "exp3", "exp4", "exp5"), 
                         conthits = T, aicc = F, bidirectional=F, bmd_low_bnd = 0.4)
```

```{r example even lower bound-res}
# print out the new results
# add to previous results for comparison 
res_low3['Min. Conc.'] <- min(conc2)
res_low3['Name'] <- paste("Chlorothalonil", "after `censoring` (two fifths)", sep = "-")
res_low3[1, c("Min. Conc.", "bmd", "bmdl", "bmdu")] <- round(res_low3[1, c("Min. Conc.", "bmd", "bmdl", "bmdu")], 3)

output_low <- rbind(output_low[-3, ], 
                    res_low3[1, c('Name', "Min. Conc.", "bmd", "bmdl", "bmdu")])
DT::datatable(output_low,rownames = FALSE)
```

```{r example even lower bound-plot, class.source="scroll-100"}
# initiate the plot
plot(conc2,resp2,xlab="conc (uM)",ylab="Response",xlim=c(0.001,100),ylim=c(-5,60),
       log="x",main=paste(name,"\n",assay),cex.main=0.9)

# add vertical lines to mark the minimum concentration in the data and the lower threshold set by bmd_low_bnd
abline(v=min(conc2), lty = 2, col = "brown", lwd = 2)
abline(v=0.4*min(conc2), lty = 2, col = "darkviolet", lwd = 2)

# add markers for BMD and its boundaries before `censoring`
lines(c(res_low$bmd,res_low$bmd),c(0,50),col="green",lwd=2)
rect(xleft=res_low$bmdl,ybottom=0,xright=res_low$bmdu,ytop=50,col=rgb(0,1,0, alpha = .5), border = NA)
points(res_low$bmd, 0, pch = "x", col = "green")

# add markers for BMD and its boundaries after `censoring`
lines(c(res_low3$bmd,res_low3$bmd),c(0,50),col="blue",lwd=2)
rect(xleft=res_low3$bmdl,ybottom=0,xright=res_low3$bmdu,ytop=50,col=rgb(0,0,1, alpha = .5), border = NA)
points(res_low3$bmd, 0, pch = "x", col = "blue")

# add the fitted curve
lines(conc_plot, exp4(ps = c(res_low$tp, res_low$ga), conc_plot))
```

**Figure 8**: This plot shows how BMD and its boundaries changed after "censoring" with a lower threshold value: 40% of 0.6, which is 0.24. The violet dash line represents the minimum concentration in the data, and the brown dash line represents the threshold value set by `bmd_low_bnd`. BMD = 0.302 falls between the lowest concentration and the threshold, and it will remain unchanged. A green column and a green "X" are supposed to mark the original BMD estimate and it's boundaries, and a blue column and a blue "X" are supposed to mark the "censored" BMD and its boundaries. In this case, they are completely overlapped. 

