---
title: "tcplfit2 Vignette"
author: "Center for Computational Toxicology and Exposure"
output:
   prettydoc::html_pretty:
    theme: architect
    toc: yes
    toc_depth: 4
vignette: >
  %\VignetteIndexEntry{tcplfit2_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


```{r setup, warning = FALSE, echo = FALSE}
library(tcplfit2)
```
# Getting started with tcplfit2
The package tcplfit2 contains the core concentration-response functionality of the package tcpl (The ToxCast Pipeline) built to process all of the ToxCast high-throughput screen (HTS) data at the US EPA. Much of the rest of the code in tcpl is used to do data processing, normalization, and database storage. We wanted to reuse the core concentration-response code for other projects, and add extensions to it, which was the origin of the current package `tcplfit2`. The main set of extensions was to include all of the concentration-response models that are contained in the program [BMDExpress](https://www.sciome.com/bmdexpress/). 
These include exponential, polynomial (1 & 2), and power functions in addition to the original Hill, gain-loss and constant models.
Additionally, we wanted to include BMD (Benchmark Dose Modeling) outputs, which is simply defining a 
Benchmark Response (BMR) level and setting the BMD to the concentration where the curve crosses the BMR level. 
One final addition was to let the hitcall value be a continuous number ranging from 0 to 1. Continuous hitcall in `tcplfit2` are defined as the product of three proportional weights: 1) the AIC of the winning model is better than the constant model (i.e. winning model is not fit to background noise), 2) at least one concentration has a median response that exceeds cutoff, and 3) the top from the winning model exceeds the cutoff. This vignette describes some functionality of the `tcplfit2` package with a few simple examples.

## Example 1: Running a single concentration-response calculation
All calculations use the function `concRespCore` which has several key inputs. The first set are 
put into a named list called 'row':

* `conc` - a vector of concentrations (not log concentrations)
* `resp` - a vector of responses, of the same length as `conc`. Note that replicates are allowed, i.e. there can be multiple pairs of conc and resp with the same concentration value.
* `cutoff`- this is the value that the response must exceed before a a curve can be called a hit. For ToxCast, this is usually some multiple (typically 3) of the median absolute deviation (BMAD) around baseline for the lowest two concentration. The user is free to make other choices
* `bmed` - this is the median of the baseline. The entire response series will be shifted by this amount. Set to zero if the data is zero-centered.  
* `onesd`- This is one standard deviation of the noise around the baseline. The BMR value = `onesd`*`bmr_scale`. The default `bmr_scale` is 1.349.

The function `concRespCore` can also have other optional elements which will be included in the output. These can be,
for instance, the name of the chemical (or other identifiers) or the name of the assay being modeled. Two other 
parameters might be used. The first is a Boolean `conthits`. If TRUE (the default, and recommended usage), the hitcall
returned will be a continuous value between 0 and 1. The other is `do.plot`. If this is set to TRUE (default is FALSE),
a plot of the curve will be generated. The user can also select only a subset of the models to be run. The example below has all of the possible ones
included. If the `fitmodels` parameter is specified, then it must always include the constant model (`cnst`) since this provides one basis for comparison during hitcalling.  However, some models may be excluded, for example the gain-loss (`gnls`) model is excluded in some applications.

To run a simple example, use the following code ...
```{r example1, fig.height = 4.55, fig.width = 8}
  conc <- list(.03,.1,.3,1,3,10,30,100)
  resp <- list(0,.2,.1,.4,.7,.9,.6, 1.2)
  row = list(conc = conc, resp = resp, bmed = 0, cutoff = 1, onesd = .5,name="some chemical")
  oldpar <- par(no.readonly = TRUE)
  on.exit(par(oldpar))   
  par(xpd = TRUE)
  res <- concRespCore(row,fitmodels = c("cnst", "hill", "gnls", "poly1", "poly2", "pow", "exp2", "exp3",
                                        "exp4", "exp5"),conthits = T, do.plot=T)
  
```

The output of this run will be a data frame with one row, summarizing the results for the winning model.

```{r example1 result, warning=FALSE, echo=FALSE}
library(DT)
DT::datatable(res,rownames = FALSE,options = list(scrollX = T))
```

## Example 2: Running a series of concentration-response models for a single assay
The input data for this example is taken from one of the Tox21 HTS assays, for estrogen receptor (ER) 
agonist activity. The data is from the mc3 table in the database `invitrodb`, which is the back end for
`tcpl`. The input data for this example have already been through the pre-processing steps (prior to `tcpl`), and processed through several steps in the ToxCast pipeline (including normalization of concentration data and transformation of response values - Levels 0 - 3). This example will run 6 chemicals out of the 100 that are included in the data set, and will create plots 
for these. The plotting routine `concRespPlot` is somewhat generic, and we anticipate that users will make
their own version of this. To run this example, use the following code ...
```{r example2, fig.height = 8, fig.width = 7}
  # read in the data
  # Loading in the level 3 example data set from invitrodb
  data("mc3")

  # set up a 3 x 2 grid for the plots
  oldpar <- par(no.readonly = TRUE)
  on.exit(par(oldpar))            
  par(mfrow=c(3,2),mar=c(4,4,2,2))

  # determine the background variation
  temp <- mc3[mc3$logc<= -2,"resp"]
  bmad <- mad(temp)
  onesd <- sd(temp)
  cutoff <- 3*bmad

  # select six samples. Note that there may be more than one sample processed for a given chemical
  spid.list <- unique(mc3$spid)
  spid.list <- spid.list[1:6]

  for(spid in spid.list) {
    # select the data for just this sample
    temp <- mc3[is.element(mc3$spid,spid),]

    # The data file has stored concentration in log10 form, so fix that
    conc <- 10**temp$logc
    resp <- temp$resp

    # pull out all of the chemical identifiers and the name of the assay
    dtxsid <- temp[1,"dtxsid"]
    casrn <- temp[1,"casrn"]
    name <- temp[1,"name"]
    assay <- temp[1,"assay"]

    # create the row object
    row <- list(conc = conc, resp = resp, bmed = 0, cutoff = cutoff, onesd = onesd,assay=assay,dtxsid=dtxsid,casrn=casrn,name=name)

    # run the concentration-response modeling for a single sample
    res <- concRespCore(row,fitmodels = c("cnst", "hill", "gnls", "poly1", "poly2", "pow", "exp2", "exp3",
                                          "exp4", "exp5"),conthits = T, aicc = F,bidirectional=F)

    # plot the results
    concRespPlot(res,ymin=-10,ymax=100)
  }


```

One would typically save the result rows in a data frame end export these for further analysis. You could remove the plotting function from the current loop and have a loop that read from the overall results data frame and only plot selected results (e.g. those with significant responses).

```{r example2 result, warning=FALSE, echo=FALSE}
DT::datatable(res,rownames = FALSE,options = list(scrollX = T))
```

## Example 3: Plotting concentration-response modeling on transcriptional signatures
The input data for this example contains 6 signatures for one chemical in a transcriptomics data set. Each signature is a different assay endpoint, thus one row in the data represents a given chemical and signature pair (assay endpoint). This data set is a sample from the signature scoring method that provides the cutoff, one standard deviation, and the concentration-response data. The example illustrates two kinds of plots available in `tcplfit2`. In the call to `concRespCore()`, the argument `do.plot` is set to `TRUE`, which provides a simple plot showing results of all the different curve fitting methods. Next, utilizing the function `concRespPlot()` provides a more informative plot for the winning model.
```{r example3, fig.height = 6, fig.width = 7, warning = FALSE}
  # call additional R packages
  library(stringr)  # string management package

  # read in the file
  data("signatures")
  
  # set up a 3 x 2 grid for the plots
  oldpar <- par(no.readonly = TRUE)
  on.exit(par(oldpar))            
  par(mfrow=c(3,2),mar=c(4,4,2,2))
    
  # fit 6 observations in signatures
  for(i in 1:nrow(signatures)){
    # set up input data
    row = list(conc=as.numeric(str_split(signatures[i,"conc"],"\\|")[[1]]),
               resp=as.numeric(str_split(signatures[i,"resp"],"\\|")[[1]]),
               bmed=0,
               cutoff=signatures[i,"cutoff"],
               onesd=signatures[i,"onesd"],
               name=signatures[i,"name"],
               assay=signatures[i,"signature"])
    # run concentration-response modeling (1st plotting option)
    out = concRespCore(row,conthits=F,do.plot=T)
    if(i==1){
      res <- out
    }else{
      res <- rbind.data.frame(res,out)
    }
  }
```

```{r example3_plot2, fig.height = 8, fig.width = 7}
  # set up a 3 x 2 grid for the plots
  oldpar <- par(no.readonly = TRUE)
  on.exit(par(oldpar))            
  par(mfrow=c(3,2),mar=c(4,4,2,2))
  # plot results using `concRespPlot`(2nd plotting option)
  for(i in 1:nrow(res)){
    concRespPlot(res[i,],ymin=-1,ymax=1)
  }
```

## Example 4: Running tcpl-like multi-concentration response data without a database connection
The ToxCast pipeline `tcpl` is an R package that manages, curve-fits, plots, and stores ToxCast data to populate its linked MySQL database, InvitroDB. The original `tcplFit()` function within `tcpl` performed basic concentration response curve fitting. Processing with tcpl_v3 and beyond depends on `tcplfit2` to allow a wider variety of concentration-response models when using `invitrodb` in the 4.0 schema and beyond. `tcplLite` was deprecated with the updates to `tcpl` and development of `tcplfit2`, since `tcplfit2` allows one to perform curve-fitting and hit-calling independent of a database. The example below demonstrates how to perform an analogous `tcplLite` analysis with `tcplfit2`. For additional information, please consult vignettes for `library(tcpl)` at https://CRAN.R-project.org/package=tcpl. 

The input for this example comes from the ACEA_AR assay. Data from the assay component ACEA_AR_agonist_80hr was analyzed in the positive analysis fitting direction relative to DMSO as the neutral control and baseline of activity. Using a electrical impedance as a cell growth reporter, increased activity can be used to infer increased signaling at the pathway-level for the androgen receptor (as encoded by the AR gene). Given heterogeneous assay data, source data often must go through pre-processing steps to transform into a uniform data format, often like this level 0. The below table is identical to the multi-concentration level 0 data (mc0) table that would be seen in `invitrodb` and recognized by `tcpl`. Columns include:

* m0id = Level 0 id
* spid = Sample id
* acid = Unique assay component id; unique numeric id for each assay component
* apid = Assay plate id
* coli = Column index (location on assay plate)
* rowi = Row index (location on assay plate)
* wllt = well type
* wllq = well quality
* conc = concentration
* rval = raw value
* srcf = Source file name
* clowder_uid = clowder unique id for source files
* git_hash = hash key for pre-processing scripts

```{r example4_init, fig.height = 6, fig.width = 7, message=FALSE, warning = FALSE}
# Loading in the level 0 example data set from invitrodb
data("mc0")
library(data.table)
dat <- mc0
DT::datatable(head(dat[wllt=='t',]),rownames= FALSE, options = list(scrollX = T))
```

To run standalone `tcplfit2` fitting without the need for a MySQL database connection like `invitrodb`, the user will replicate stepping through the multiple levels of processing. A detailed explanation of processing levels can be found within `tcpl`'s Data Processing vignette.

Level 1 importantly establishes the concentration index. The concentration index is simply the distinct concentrations ranked from lowest to highest, and this index can be used to calculate the baseline median absolute deviation for an assay.

```{r example4_cndx, fig.height = 6, fig.width = 7}
library(tcpl)
## Order by the following columns
setkeyv(dat, c('acid', 'srcf', 'apid', 'coli', 'rowi', 'spid', 'conc'))

## Define replicate id (rpid) column for test compound wells
nconc <- dat[wllt == "t" , ## denotes test well as the well type (wllt)
             list(n = lu(conc)), #total number of unique concentrations
             by = list(acid, apid, spid)][ , list(nconc = min(n)), by = acid]
dat[wllt == "t" & acid %in% nconc[nconc > 1, acid],
    rpid := paste(acid, spid, wllt, srcf, apid, "rep1", conc, sep = "_")]
dat[wllt == "t" & acid %in% nconc[nconc == 1, acid],
    rpid := paste(acid, spid, wllt, srcf, "rep1", conc, sep = "_")]

## Define rpid column for non-test compound wells
dat[wllt != "t",
    rpid := paste(acid, spid, wllt, srcf, apid, "rep1", conc, sep = "_")]

## set repid based on rowid
dat[, dat_rpid := rowid(rpid)]
dat[, rpid := sub("_rep[0-9]+.*", "",rpid, useBytes = TRUE)]
dat[, rpid := paste0(rpid,"_rep",dat_rpid)]

# Define concentration index
indexfunc <- function(x) as.integer(rank(unique(x))[match(x, unique(x))])
dat[ , cndx := indexfunc(conc), by = list(rpid)]
```

### Adjustments
Levels 2 and 3 are used for data adjustments and normalization. Generally if the response values (`rval`) need to be logged or transformed in some way from their original values this is where that adjustment would occur. Transformed response values are referred to as corrected values and are stored in the `cval` field/variable. However, in this case, the corrected values (`cval`) are identical to the original response values (`rval`). 

```{r example4_mc2, fig.height = 6, fig.width = 7}
# If no adjustments are required for the data, the corrected value (cval) should be set as original rval
dat[,cval := rval]

## Poor well quality (wllq) wells should be removed
dat <- dat[!wllq == 0,]

## Fitting generally cannot occur if response values are NA therefore values need to be removed
dat <- dat[!is.na(cval),]

## A column for log10 concentration is added as some of the mc3 methods require logc. Given logging concentration, conc=0 are not allowed therefore a dummy non-zero value should be used
dat[conc == 0 , conc := 0.0001]
dat[ , logc := log10(conc)]

#As a final step to prepare the dataset tcplfit2 processing, a dummy aeid is required if using mc3_mthds from tcpl
dummy_aeid <- 99999
dat[,aeid := dummy_aeid]

## Set aeid as a key
setkey(dat,aeid)
```

Once the data is initialized to a point where the required fields are available, the methods included in the `tcpl` package can be identified and applied without the need for a database connection. You can see the list of available methods for Level 3 in the table below:

```{r example4_mthdlist, fig.height = 6, fig.width = 7, warning = FALSE}
mthd_funcs <- tcpl:::mc3_mthds()
DT::datatable(tcpl::tcplMthdList(3),rownames= FALSE, options = list(scrollX = T))
```

### Normalization
Here three normalization methods are selected and applied to the data.  Note because of the way `tcpl` handles the application of functions, the dataframe must be called `dat`. In the future, `tcpl` will export these functions so that they can be applied to any dataset without the need for a specific name or dummy aeid.

```{r example4_mc3methods, fig.height = 6, fig.width = 7, results = 'hide'}
# apply level 3 methods
## These methods directly apply the normalization methods from tcpl without the need for a DB connection
lapply(mthd_funcs[["bval.apid.nwlls.med"]](dummy_aeid), eval)
lapply(mthd_funcs[["pval.apid.medncbyconc.min"]](dummy_aeid),eval)
lapply(mthd_funcs[["resp.pc"]](dummy_aeid),eval)
```

Level 4 determines the baseline variability, or noise, that will later be used for cutoff calculation. Using the established concentration index, the level 4 methods can be loaded in a similar way to level 3.

```{r example4_mthdlist_l4, fig.height = 6, fig.width = 7}
mthd_funcs_l4 <- tcpl:::mc4_mthds()
DT::datatable(tcpl::tcplMthdList(4), rownames= FALSE, options = list(scrollX = T))
```

There are much fewer level 4 methods, but generally it is a requirement to assign a method that calculates the bmad and assign a method that calculates the standard deviation of the noise for `tcplfit2` fitting.

```{r example4_mc4methods, fig.height = 6, fig.width = 7, results = 'hide'}
# apply level 4 methods
## These methods directly apply the noise calculation and fitting methods from tcpl without the need for a DB connection
lapply(mthd_funcs_l4[["bmad.aeid.lowconc.twells"]](),eval)
lapply(mthd_funcs_l4[["onesd.aeid.lowconc.twells"]](),eval)
lapply(mthd_funcs_l4[["bidirectional.false"]](),eval)
```

### Dose-Response Curve Fitting
After methods up to level 4 have been applied, the model fitting can begin. In `tcpl`, this would be considered level 4, and is where `tcplfit2` is used to fit all of the models as a dependency for `tcpl`.

```{r example4_fitting, fig.height = 6, fig.width = 7}
#do tcplfit2 fitting
myfun <- function(y) {
  res <- tcplfit2::tcplfit2_core(y$conc,
                          y$resp,
                          cutoff = unique(y$bmad),
                          bidirectional = TRUE,
                          verbose = FALSE,
                          force.fit = TRUE,
                          fitmodels = c("cnst", "hill", "gnls", "poly1",
                                        "poly2", "pow", "exp2", "exp3",
                                        "exp4", "exp5")
                          )
  list(list(res)) #use list twice because data.table uses list(.) to look for values to assign to columns
}
```

The following code performs dose-response modeling for all spids in the dataset. **Warning: The fitting step for the full data set, `dat`, can take 7-10 minutes to run.** Hence the code chunk following provides a subset example of data for curve fitting and hitcalling. The subset data only contains records of six samples.

```{r example4_fitting_full, eval=FALSE}
# only want to run tcplfit2 for test wells in this case
# this chunk doesn't run, fit the curves on the subset below
dat[wllt == 't',params:= myfun(.SD), by = .(spid)]
```

```{r example4_fitting_subset}
# create a subset that contains 6 samples and run curve fitting
subdat <- dat[spid %in% unique(spid)[10:15],]
subdat[wllt == 't',params:= myfun(.SD), by = .(spid)]
```

### Continuous Hitcalling
After all of the models have been fit, hitcalling can occur. The output of level 4 can be fed directly into the `tcplhit2_core` function. The results are then pivoted and shown in the resulting datatable.

```{r example4_hitcalling, fig.height = 6, fig.width = 7}
myfun2 <- function(y) {
  res <- tcplfit2::tcplhit2_core(params = y$params[[1]],
                                 conc = y$conc,
                                 resp = y$resp,
                                 cutoff = 3*unique(y$bmad),
                                 onesd = unique(y$osd)
                                 )
  list(list(res))
}

# continute with hitcalling
res <- subdat[wllt == 't', myfun2(.SD), by = .(spid)]

#pivot wider
res_wide <- rbindlist(Map(cbind, spid = res$spid, res$V1))

DT::datatable(res_wide,options = list(scrollX = T))
```

*Hitcalling can also be done with the full data set, `dat`.*

The output table resulting from the previous code chunk is the same format as the `res` table in example 3. Thus, one can use the `concRespPlot` function, as done previously in example 3, to plot the results. The next code chunk demonstrates how to visualize the example 4 fit results.

```{r example4_plot, fig.height = 8, fig.width = 7}
  # set up a 3 x 2 grid for the plots
  oldpar <- par(no.readonly = TRUE)
  on.exit(par(oldpar))            
  par(mfrow=c(3,2),mar=c(4,4,2,2))
  # plot results using `concRespPlot`(2nd plotting option)
  for(i in 1:nrow(res)){
    concRespPlot(res_wide[i,],ymin=-50,ymax=50)
  }
```

## Example 5: How can you censor BMD
When doing curve fitting, BMD is returned for curve with hit-call > 0. Users have the option to "censor" the BMD when it occurs above or below an experimental dose or other specified values. This section provides examples of how to do it in situations of when the BMD is lower or higher than the minimum or maximum concentration in the data with arguments `bmd_low_bnd` and `bmd_up_bnd`. 

First, let's look at situation one: when BMD is below the lowest concentration tested. The following code chunk loads data of a sample from `mc3` and removes data in some of the low dose groups so it simulates a curve that takes a huge lift in the first couple dose groups. In this situation, the BMD is more likely to be a lower dose that are not in the data. 

```{r example 5 lower}
# load example data
spid <- unique(mc3$spid)[94]
temp <- mc3[is.element(mc3$spid,spid),]

# The data file has stored concentration in log10 form, fix it 
conc <- 10**temp$logc # back-transforming concentrations on log10 scale
resp <- temp$resp

# modify the data for demonstration purposes 
conc2 <- conc[conc>0.41]
resp2 <- resp[which(conc>0.41)]

# pull out all of the chemical identifiers and the name of the assay
dtxsid <- temp[1,"dtxsid"]
casrn <- temp[1,"casrn"]
name <- temp[1,"name"]
assay <- temp[1,"assay"]

# create the row object
row <- list(conc = conc2, resp = resp2, bmed = 0, cutoff = cutoff, onesd = onesd,assay=assay,dtxsid=dtxsid,casrn=casrn,name=name)

# run the concentration-response modeling for a single sample
res_low <- concRespCore(row,fitmodels = c("cnst", "hill", "gnls", "poly1", "poly2", "pow", "exp2", "exp3", "exp4", "exp5"), conthits = T, aicc = F, bidirectional=F)

concRespPlot(res_low,ymin=-10,ymax=100)
```

```{r example 5 lower-res, echo=FALSE}
# function results
res_low['Min. Conc.'] <- min(conc2)
DT::datatable(round(res_low[1, c("Min. Conc.", "bmd", "bmdl", "bmdu")], 3),rownames = FALSE)
```

The lowest concentration tested is 0.6 but the estimated BMD from the function is 0.302, it is lower than the minimum concentration tested. This could happen due to experimental design or it implies the analysis is unreliable.

Users can allow BMD to be lower than the minimum experimental dose range while restrict it to be no lower than a percentage of the lowest concentration by using the argument `bmd_low_bnd`. This is a multiplier that times with the lowest dose in the data to generate a allowable lower boundary for BMD. If set the argument to 0.8, we don't want the BMD to be lower than 80% of the lowest concentration tested (0.8*0.6 = 0.48). If the BMD is not lower than this allowable lower boundary, than it will be returned as-is. However, in this example 0.302 is still lower than 0.48, so BMD will be forced to be the lowest allowable value and the original BMD boundaries in the output from `concRespCore` are also being shifted.

```{r example 5 lower-demo}
# using the argument to set a lower bound for BMD
res_low2 <- concRespCore(row,fitmodels = c("cnst", "hill", "gnls", "poly1", "poly2", "pow", "exp2", "exp3", "exp4", "exp5"), conthits = T, aicc = F, bidirectional=F, bmd_low_bnd = 0.8)
```

```{r example 5 new lower-res, echo=FALSE}
# print out the new results
res_low2['Min. Conc.'] <- min(conc2)
DT::datatable(round(res_low2[1, c('Min. Conc.',"bmd", "bmdl", "bmdu")],3),rownames = FALSE)
```

```{r example 5 lower cont.}
logc_plot <- seq(from=-3,to=2,by=0.05)
conc_plot <- 10**logc_plot

plot(c(1,1),type="n",xlab="conc (uM)",ylab="Response",xlim=c(0.001,100),ylim=c(0,100),
       log="x",main=paste(name,"\n",assay),cex.main=0.9)

lines(c(res_low$bmd,res_low$bmd),c(0,50),col="green",lwd=2)
rect(xleft=res_low$bmdl,ybottom=0,xright=res_low$bmdu,ytop=50,col=rgb(0,1,0, alpha = .5), border = NA)

lines(c(res_low2$bmd,res_low2$bmd),c(0,50),col="red",lwd=2)
rect(xleft=res_low2$bmdl,ybottom=0,xright=res_low2$bmdu,ytop=50,col=rgb(1,0,0, alpha = .5), border = NA)

points(res_low$bmd, 0, pch = "x", col = "blue")
points(res_low2$bmd, 0, pch = "x", col = "red")
points(conc2, resp2)
lines(conc_plot, exp4(ps = c(res_low$tp, res_low$ga), conc_plot))
```

**Figure x**: This plot shows how BMD and its boundaries changes after "censoring". The red line is the BMD after "censoring" and the red column is the updated boundaries. The BMD and both boundaries were shifted right and the boundaries are narrower.  

`bmd_up_bnd` is also a multiplier that times with the maximum concentration in the data to generate another allowable boundary. 

```{r example 5 upper}
# load example data
spid <- unique(mc3$spid)[26]
temp <- mc3[is.element(mc3$spid,spid),]

# The data file has stored concentration in log10 form, so fix that
conc <- 10**temp$logc # back-transforming concentrations on log10 scale
resp <- temp$resp

# pull out all of the chemical identifiers and the name of the assay
dtxsid <- temp[1,"dtxsid"]
casrn <- temp[1,"casrn"]
name <- temp[1,"name"]
assay <- temp[1,"assay"]

# create the row object
row <- list(conc = conc, resp = resp, bmed = 0, cutoff = cutoff, onesd = onesd,assay=assay,dtxsid=dtxsid,casrn=casrn,name=name)

# run the concentration-response modeling for a single sample
res_up <- concRespCore(row,fitmodels = c("cnst", "hill", "gnls", "poly1", "poly2", "pow", "exp2", "exp3", "exp4", "exp5"), conthits = T, aicc = F, bidirectional=F)

concRespPlot(res_up,ymin=-10,ymax=100)
```

```{r example 5 upper-res, echo=FALSE}
# max conc.
res_up['Max Conc.'] <- max(conc)
# function results
DT::datatable(round(res_up[1, c('Max Conc.', "bmd", "bmdl", "bmdu")],3),rownames = FALSE)
```

The BMD 300 is higher than the maximum concentration tested, 80. 

The documentation use 10 as example, but I think there are two  from here: 

If we input a value that's larger than 1, that means we allow BMD to be larger than the maximum concentration tested, but no large than a multiple of the maximum concentration. If BMD is still larger than the multiple of maximum concentration, it's forced to be the maximum allowable value. 

If we input a decimal, that means we want BMD to be no larger than a percentage of the maximum concentration in the data (if it's not 1, it prevents the maximum dose to be the BMD). 

```{r example upper-2}
# input a number larger than 1
res_up2 <- concRespCore(row,fitmodels = c("cnst", "hill", "gnls", "poly1", "poly2", "pow", "exp2", "exp3", "exp4", "exp5"), conthits = T, aicc = F, bidirectional=F, bmd_up_bnd = 2)
```

```{r example 5 upper cont., echo=FALSE}
res_up2['Max Conc.'] <- max(conc)
DT::datatable(res_up2[1, c('Max Conc.',"bmd", "bmdl", "bmdu")],rownames = FALSE,options = list(scrollX = T))
```

# input a decimal 
```{r}
res_up3 <- concRespCore(row,fitmodels = c("cnst", "hill", "gnls", "poly1", "poly2", "pow", "exp2", "exp3", "exp4", "exp5"), conthits = T, aicc = F, bidirectional=F, bmd_up_bnd = 0.5)
```

```{r echo=FALSE}
res_up3['Max Conc.'] <- max(conc)
DT::datatable(res_up3[1, c('Max Conc.', "bmd", "bmdl", "bmdu")],rownames = FALSE,options = list(scrollX = T))
```


